---
title: "1. Fast Intro to R"
subtitle: "307102 Descriptive Statistics for Business | University of Petra"

# Two outputs from this one file: the notebook page, and the lecture deck.
#
# Content that should differ between them is wrapped in a div:
#   ::: {.content-visible unless-format="revealjs"}   notebook prose
#   ::: {.content-visible when-format="revealjs"}     slide version
#
# NOTE: do NOT use when-format="html" for the notebook half. Revealjs is an
# HTML-derived format, so it matches "html" too and BOTH versions appear on
# the slides. Always pair when-format="revealjs" with unless-format="revealjs".
format:
  html: default
  revealjs:
    output-file: Intro-to-R-slides.html
    theme: [default, ../assets/petra-reveal.scss]
    width: 1280
    height: 800
    slide-number: c/t
    show-slide-number: all
    scrollable: true
    chalkboard: true
    preview-links: auto
    history: false
    footer: "307102 · Section 1 · Intro to R · University of Petra"
---

This is the only part of the course that is about R rather than about statistics.
It exists so that everything afterwards can be about statistics.

There are **eight things** to learn here. That is the whole list. You will not be
asked to write loops, define your own functions, or do anything a programmer would
recognise as programming. You are learning to operate a tool, not to build one.

Expect this to take about ninety minutes. Do not skip it - every later notebook
assumes you have done it.

::: {.callout-tip title="Before you start"}
You will make typing mistakes and R will respond with error messages. This happens
to everyone, including people who have used R for twenty years.

An error is not a sign that something has gone seriously wrong. It is R saying "I
did not understand that", which is almost always a missing bracket, a missing quote,
or a misspelled name. Read the message, check those three things, and try again.
:::

::: {.content-visible unless-format="revealjs"}
::: {.callout-tip title="Lecture slides"}
This same material is also available as a slide deck:
[**Open the slides →**](Intro-to-R-slides.html)

Arrow keys to move, **F** for full screen, **S** for the presenter view with
speaker notes, **B** for the chalkboard, **Esc** for a grid of all slides.
:::
:::

## Objectives {.smaller}
By the end of this notebook you will be able to:

- Run R code and read what R prints back to you
- Store a value in an object and use it later
- Explain what a data frame is and why it matters
- Load a spreadsheet into R
- Pull a single column out of a dataset
- Read code written with the pipe operator
- Use five commands to filter, sort, summarise and reshape data

## How to use this notebook

This file is an **R Notebook**. It mixes explanation (like this paragraph) with
**code chunks** (the grey boxes).

To run a chunk, click the small green triangle at its top-right corner, or click
inside it and press **Ctrl+Shift+Enter** (**Cmd+Shift+Enter** on Mac). The result
appears directly beneath.

Run every chunk as you come to it. Reading code without running it teaches you
almost nothing.

# 1. R is a calculator that remembers

At its simplest, R is a calculator.

```{r calculator}
2 + 2

150 * 0.05

(45 + 52 + 38) / 3
```

The `[1]` in front of each answer is not part of the answer. It is R numbering the
output, telling you this line begins with the first result. You will see it
constantly. Ignore it.

So far this is Excel with extra steps. What makes R different is the next idea.

# 2. Objects - giving things names

In Excel a number lives in a cell and you refer to it by address, like `B7`. Insert
a row and `B7` becomes something else, and your formulas break.

In R you give a value a **name** instead. The name never moves.

The arrow `<-` means "store this under this name". Read it as "gets".

```{r objects}
tax_rate <- 0.05

item_price <- 120

tax_amount <- item_price * tax_rate

tax_amount
```

Three things worth noticing.

**Storing something produces no output.** When you ran `tax_rate <- 0.05`, nothing
printed. That is correct - you told R to remember a value, not to show it. To see
what is stored, type the name on its own, as the last line does.

**Names should say what the thing is.** `tax_amount` is a good name. `x` is a bad
one. In three weeks you will not remember what `x` was.

**Names are case-sensitive and cannot contain spaces.** `Tax_Rate` and `tax_rate`
are different names. Use lowercase with underscores and you will avoid most
problems.

Once a value is stored you can reuse it as often as you like:

```{r objects_reuse}
total_paid <- item_price + tax_amount
total_paid
```

Look at the **Environment** pane in the top right of RStudio. Every name you have
created is listed there. That pane is your R memory - if something is not in it, R
does not know about it.

# 3. Functions - verbs that do things

A **function** is a command that does something to a value. You have used functions
in Excel: `SUM`, `AVERAGE`, `COUNT`. R works the same way with different names.

A function is its name followed by round brackets, with the thing it works on
inside.

```{r functions_intro}
prices <- c(12.5, 8.0, 45.2, 19.9, 33.1)

mean(prices)

max(prices)

length(prices)
```

The `c()` is itself a function. It stands for "combine" and glues several values
into a single list of numbers. In R this is a **vector** - simply a column of values
of the same type.

Some functions take extra instructions, called **arguments**, separated by commas:

```{r functions_arguments}
messy_prices <- c(12.5, 8.0, NA, 19.9, 33.1)

mean(messy_prices)

mean(messy_prices, na.rm = TRUE)
```

`NA` means "missing value" - a blank cell. The first `mean()` returns `NA` because R
refuses to guess: if one value is unknown, the true average is unknown too. The
argument `na.rm = TRUE` says "remove the missing values first, then average what is
left".

::: {.callout-warning title="Remember this one"}
When a calculation unexpectedly returns `NA`, **missing data is almost always the
reason**. This causes more confusion later in the course than any other single
thing.
:::

:::: {.callout-note title="Self-check 1"}
Without running it, what will `length(messy_prices)` return - 4 or 5?

::: {.callout-note collapse="true" title="Show the answer"}
**5.** `length()` counts how many slots the vector has, and the missing value still
occupies a slot. `NA` means "there is a value here and we do not know it", not
"there is nothing here".

That distinction matters commercially: a customer who declined to give their age is
not the same as a customer who does not exist.
:::
::::

# 4. Data frames - the spreadsheet inside R

A single vector is one column. Real data has many columns of different types - some
numbers, some text, some dates - all lined up so that row 7 in every column refers
to the same transaction.

That structure is a **data frame**. It is the single most important idea in this
notebook: a spreadsheet that lives inside R.

Here is a tiny one, built by hand:

```{r data_frame_manual}
library(tidyverse)

branches <- data.frame(
  branch      = c("A", "B", "C"),
  city        = c("Yangon", "Mandalay", "Naypyitaw"),
  staff_count = c(24, 19, 21),
  open_year   = c(2015, 2017, 2016)
)

branches
```

Three rows, four columns. Each **row** is one branch. Each **column** is one piece
of information recorded about every branch.

::: {.callout-tip title="The rule that makes analysis possible"}
**One row per thing, one column per measurement.**

Almost everything you do in this course assumes your data is arranged this way.
When an analysis will not work, the first thing to check is whether the data
actually follows this rule.
:::

::: {.content-visible when-format="revealjs"}
::: {.notes}
This is the idea to slow down on. Almost every problem students bring to office
hours is data that is not shaped this way. Ask them to say out loud what one row
represents; if they cannot, that is the bug.
:::
:::

The `library(tidyverse)` line loads the add-on packages you installed during setup.
You need one `library()` call at the start of every session. If you restart R, run
it again.

# 5. Loading real data

You will not usually type data by hand. You will load it from a file.

For the rest of this course the file is `supermarket_sales.xlsx`: 1,000 transactions
from a supermarket chain with three branches. `read_excel()` reads it.

```{r load_data}
library(readxl)

sales <- read_excel("../data/supermarket_sales.xlsx")

sales
```

The path `"../data/supermarket_sales.xlsx"` reads as: go up one folder from where
this notebook lives (`..`), then into `data`, then take that file.

If you get an error saying the file does not exist, open `307102-R-Track.Rproj` in
the `R-Version` folder and work from there. That tells RStudio which folder is
"home".

## Getting your bearings

Before analysing any dataset, look at it. Three functions do this.

```{r inspect_dim}
dim(sales)
```

`dim()` gives the size: rows first, then columns. So 1,000 transactions, each with
17 pieces of information.

```{r inspect_names}
names(sales)
```

`names()` lists the column names. Notice that several contain spaces - `Unit price`,
`Product line`, `Customer type`. That is common in real business data and it has a
consequence you will meet in a moment.

```{r inspect_glimpse}
glimpse(sales)
```

`glimpse()` is the most useful of the three. Each line is one column, showing its
name, its **type**, and the first few values.

Those types matter:

| What you see | What it means | Example column |
|---|---|---|
| `<chr>` | Character - text | `Branch`, `Gender` |
| `<dbl>` | Double - a number that can have decimals | `Total`, `Rating` |
| `<int>` | Integer - a whole number | `Quantity` |
| `<dttm>` | Date and time | `Date` |

R will refuse to average a `<chr>` column, and it is right to. The average of "Male"
and "Female" is not a meaningful quantity.

But R will **not** protect you from averaging something numeric that should not be
averaged. If a variable were coded 1 = Cash, 2 = Credit, 3 = Ewallet, R would
happily report a mean of 2.03. Only you can know that number is meaningless.

# 6. The dollar sign - pulling out one column

To work with a single column, write the data frame name, then `$`, then the column
name.

```{r dollar_basic}
mean(sales$Rating)
```

Read `sales$Rating` as "the Rating column of sales".

Because `sales$Rating` is just a vector of numbers, every function from section 3
works on it:

```{r dollar_more}
max(sales$Total)

min(sales$Total)

length(sales$Total)
```

::: {.callout-warning title="Column names with spaces"}
When a column name contains a space, R needs help seeing where the name starts and
ends. Wrap it in **backticks** - the key above Tab on most keyboards.

```{r dollar_backticks}
mean(sales$`Unit price`)
```

Without backticks R reads `sales$Unit`, stops at the space, and cannot find a column
called `Unit`.

**This is the single most common error in this course**, and now you know the fix.
:::

:::: {.callout-note title="Self-check 2"}
Write one line that finds the average `gross income` per transaction.

::: {.callout-note collapse="true" title="Show the answer"}
```r
mean(sales$`gross income`)
```

Backticks are required because of the space. Note also that this column is spelled
in lowercase - `gross income`, not `Gross Income`. R is case-sensitive, so
capitalisation must match exactly what `names(sales)` showed you.
:::
::::

# 7. The pipe - reading code as a sentence

Suppose you want the average transaction value for Branch A only. Written the direct
way:

```{r pipe_nested, eval=FALSE}
summarise(filter(sales, Branch == "A"), avg = mean(Total))
```

That works, but it reads inside-out: the first thing that happens (`filter`) is
buried in the middle. With four or five steps it becomes unreadable.

The **pipe**, written `|>`, fixes this. It takes whatever is on its left and passes
it into the function on its right. It means **"and then"**.

```{r pipe_basic}
sales |>
  filter(Branch == "A") |>
  summarise(avg_transaction = mean(Total))
```

Read that top to bottom as a sentence:

> Take the sales data, **and then** keep only Branch A, **and then** compute the
> average of Total.

That is the whole idea. Every analysis in this course is written this way.

::: {.content-visible when-format="revealjs"}
::: {.notes}
Read a pipeline out loud as a sentence the first few times - "take the sales,
and then keep Branch A, and then average the total". Students who read it as
a sentence stop being frightened of it within about ten minutes.
:::
:::

Two practical points. Put the pipe at the **end** of a line, never the start of the
next one. And RStudio types it for you with **Ctrl+Shift+M** (**Cmd+Shift+M**).

# 8. The five verbs

Nearly everything you will do to a dataset is one of five operations.

| Verb | What it does | Business question it answers |
|---|---|---|
| `filter()` | Keeps some **rows** | "Show me only the credit card sales" |
| `select()` | Keeps some **columns** | "I only care about branch and total" |
| `arrange()` | **Sorts** the rows | "Which sale was largest?" |
| `mutate()` | **Adds** a new column | "What was the margin on each sale?" |
| `summarise()` | **Collapses** to a summary | "What is our average sale?" |

## filter() - choosing rows

```{r verb_filter}
sales |>
  filter(Branch == "A")
```

Note the **double** equals sign. A single `=` means "set this to"; `==` means "is
this equal to". Using `=` where you meant `==` is another very common early error.

```{r verb_filter_more}
sales |>
  filter(Total > 900)
```

Conditions can be combined. A comma means "and":

```{r verb_filter_and}
sales |>
  filter(Branch == "A", Payment == "Cash")
```

## select() - choosing columns

```{r verb_select}
sales |>
  select(Branch, `Product line`, Total, Rating)
```

Useful when a dataset has 17 columns and you care about four.

## arrange() - sorting

```{r verb_arrange}
sales |>
  arrange(desc(Total)) |>
  select(`Invoice ID`, Branch, `Product line`, Total)
```

`arrange()` sorts smallest-first by default. Wrapping the column in `desc()` sorts
largest-first, which is usually what you want when looking for top performers.

Notice this chain has two steps: sort, then narrow the columns so the result is
readable. That is the pipe earning its keep.

## mutate() - adding a column

```{r verb_mutate}
sales |>
  mutate(margin_pct = (`gross income` / Total) * 100) |>
  select(`Invoice ID`, Total, `gross income`, margin_pct)
```

`mutate()` calculates a new column from existing ones, row by row. This is the
direct equivalent of typing a formula in a new spreadsheet column and dragging it
down - except that it cannot silently fail to cover the last twenty rows.

## summarise() - collapsing to one number

```{r verb_summarise}
sales |>
  summarise(
    transactions  = n(),
    avg_sale      = mean(Total),
    total_revenue = sum(Total),
    avg_rating    = mean(Rating)
  )
```

`n()` is a helper that counts rows. It only works inside `summarise()`.

## group_by() - the one that makes summarise powerful

A single average across all 1,000 transactions is rarely the interesting number.
What you usually want is the average **for each branch**, or **for each product
line**.

```{r verb_group_by}
sales |>
  group_by(Branch) |>
  summarise(
    transactions  = n(),
    avg_sale      = mean(Total),
    total_revenue = sum(Total)
  )
```

One row per branch instead of one row overall. `group_by()` changes nothing about
the data - it only tells the next command to work separately within each group.

This is the R equivalent of a PivotTable, and it is the single most useful thing in
this notebook.

You can group by more than one column:

```{r verb_group_by_two}
sales |>
  group_by(Branch, `Customer type`) |>
  summarise(
    transactions = n(),
    avg_sale     = mean(Total),
    .groups      = "drop"
  )
```

:::: {.callout-note title="Self-check 3"}
Which product line earns the most total gross income? Write a pipeline using
`group_by()`, `summarise()` and `arrange()`.

::: {.callout-note collapse="true" title="Show the answer"}
```r
sales |>
  group_by(`Product line`) |>
  summarise(total_income = sum(`gross income`)) |>
  arrange(desc(total_income))
```

The order matters. You must group and summarise before you can sort, because the
column you are sorting by (`total_income`) does not exist until `summarise()`
creates it.

A follow-up worth thinking about: the product line with the highest total income is
not necessarily the most profitable *per sale*. Adding `mean()` of the same column
would let you check whether the leader is winning on margin or simply on volume.
That distinction is exactly what descriptive statistics exists to expose, and you
will meet it again in section 2 of the course.
:::
::::

# Putting it together {.smaller}
Here is a realistic question answered with everything above:

> Among member customers, which branch has the highest average customer rating, and
> how many transactions is that average based on?

```{r putting_together}
sales |>
  filter(`Customer type` == "Member") |>
  group_by(Branch) |>
  summarise(
    transactions = n(),
    avg_rating   = mean(Rating)
  ) |>
  arrange(desc(avg_rating))
```

Four lines, and it reads as English: take the sales, keep members only, group by
branch, work out the count and average rating for each, sort best first.

::: {.callout-tip title="Why the count column is there"}
An average of 7.2 built from 200 transactions and an average of 7.2 built from 6
transactions are not equally trustworthy.

Reporting a summary statistic without its sample size is one of the most common ways
to mislead with data. You will meet the formal version of this idea in section 3.
:::

# Your turn

Fill in the blanks. The answer to each is directly above.

**Exercise 1.** How many transactions were paid for with an Ewallet?

```{r ex1, eval=FALSE}
sales |>
  filter(Payment == "______") |>
  summarise(count = n())
```

**Exercise 2.** What is the average `Unit price` in each product line, sorted from
most expensive to least?

```{r ex2, eval=FALSE}
sales |>
  group_by(`______`) |>
  summarise(avg_price = mean(`______`)) |>
  arrange(desc(______))
```

**Exercise 3.** Add a column showing the tax as a percentage of the total, then show
only the invoice ID, total, and your new column.

```{r ex3, eval=FALSE}
sales |>
  mutate(tax_pct = (`Tax 5%` / ______) * 100) |>
  select(`Invoice ID`, ______, tax_pct)
```

**Exercise 4.** Write this one from scratch. Which single transaction had the
highest gross income, and which branch was it in?

```{r ex4, eval=FALSE}
# YOUR CODE HERE
```

::: {.callout-note collapse="true" title="Show all answers"}
```r
# Exercise 1
sales |>
  filter(Payment == "Ewallet") |>
  summarise(count = n())

# Exercise 2
sales |>
  group_by(`Product line`) |>
  summarise(avg_price = mean(`Unit price`)) |>
  arrange(desc(avg_price))

# Exercise 3
sales |>
  mutate(tax_pct = (`Tax 5%` / Total) * 100) |>
  select(`Invoice ID`, Total, tax_pct)

# Exercise 4
sales |>
  arrange(desc(`gross income`)) |>
  select(`Invoice ID`, Branch, `Product line`, `gross income`) |>
  head(1)
```

On Exercise 3, every answer comes out at 4.7619 percent rather than 5. That is not
an error. The tax is 5 percent *of the pre-tax amount*, and `Total` already includes
the tax, so the tax is a slightly smaller share of the larger number.

Noticing that a result is not quite what you expected, and then working out why, is
the habit this whole course is trying to build.
:::

# Reference card {.smaller}
Everything from this notebook in one place. Come back here whenever you are stuck.

| Code | What it does |
|---|---|
| `x <- 5` | Store 5 under the name `x` |
| `c(1, 2, 3)` | Combine values into a vector |
| `mean(x)`, `sum(x)`, `max(x)`, `min(x)` | Basic summaries |
| `length(x)` | How many values |
| `mean(x, na.rm = TRUE)` | Average, ignoring missing values |
| `library(tidyverse)` | Load the packages - once per session |
| `read_excel("path/file.xlsx")` | Load a spreadsheet |
| `dim(d)`, `names(d)`, `glimpse(d)` | Inspect a data frame |
| `d$column` | One column |
| ``d$`two words` `` | One column whose name has a space |
| `d |> f()` | Pipe: "take d, and then do f" |
| `filter(d, x == "A")` | Keep matching rows |
| `select(d, a, b)` | Keep chosen columns |
| `arrange(d, desc(x))` | Sort, largest first |
| `mutate(d, new = a / b)` | Add a calculated column |
| `summarise(d, avg = mean(x))` | Collapse to a summary |
| `group_by(d, g)` | Summarise separately per group |
| `n()` | Count rows, inside `summarise()` |

# Common errors and what they mean {.smaller}
| Message | Cause | Fix |
|---|---|---|
| `object 'sales' not found` | You have not created it, or R restarted | Run the chunk that creates it |
| `could not find function "filter"` | Packages not loaded | Run `library(tidyverse)` |
| `object 'Unit' not found` | Space in a column name | Use backticks: `` `Unit price` `` |
| `unexpected '='` | Used `=` instead of `==` | Change to `==` inside `filter()` |
| Result is `NA` | Missing values in the column | Add `na.rm = TRUE` |
| `unexpected symbol` | Missing comma, bracket or quote | Check the line *above* the one flagged |
| Console shows `+` and nothing runs | Unfinished command | Press Esc, check your brackets |

# Summary {.smaller}
You now know:

- R stores values under **names** using `<-`
- **Functions** do things and take **arguments** for extra instructions
- A **data frame** is a spreadsheet inside R: one row per thing, one column per measurement
- `read_excel()` loads data; `glimpse()` shows you what you loaded
- `$` pulls out a column, with backticks if the name has a space
- The **pipe** `|>` means "and then", letting code read like a sentence
- **Five verbs** - `filter`, `select`, `arrange`, `mutate`, `summarise` - plus `group_by()` cover most of what you need

That is the entire Fast Intro to R. Later notebooks introduce a handful of new
functions each, always with a "New R this week" table at the top, but the foundation
does not change.

**Next:** section 2, `2-Descriptive-Statistics/`, where the statistics begins.

::: {.content-visible when-format="revealjs"}
# Questions?

Work through the exercises in the notebook version of this section before
the next lecture.

::: {.notes}
Point them at the notebook rather than the slides for homework - the slides
drop the longer explanations deliberately.
:::
:::
