---
title: "3.1 Probability Distributions"
subtitle: ""

# Two outputs from this one file: the notebook page, and the lecture deck.
#
# Content that should differ between them is wrapped in a div:
#   ::: {.content-visible unless-format="revealjs"}   notebook prose
#   ::: {.content-visible when-format="revealjs"}     slide version
#
# NOTE: do NOT use when-format="html" for the notebook half. Revealjs is an
# HTML-derived format, so it matches "html" too and BOTH versions appear on
# the slides. Always pair when-format="revealjs" with unless-format="revealjs".
format:
  html: default
  revealjs:
    output-file: 3.1-Probability-Distributions-slides.html
    theme: [default, ../assets/petra-reveal.scss]
    width: 1280
    height: 800
    slide-number: c/t
    show-slide-number: all
    scrollable: true
    chalkboard: true
    preview-links: auto
    history: false
    footer: "307102 · Section 3.1 · Probability Distributions · University of Petra"
---

::: {.content-visible unless-format="revealjs"}
::: {.callout-tip title="Lecture slides"}
This same material is also available as a slide deck:
[**Open the slides →**](3.1-Probability-Distributions-slides.html)

Arrow keys to move, **F** for full screen, **S** for the presenter view with
speaker notes, **B** for the chalkboard, **Esc** for a grid of all slides.
:::
:::

## New R this week {.smaller}
R uses a single naming system for every distribution. Learn it once and you know all
of them.

| Prefix | Question it answers | Example |
|---|---|---|
| `d` | What is the probability of **exactly** this value? | `dbinom(3, 10, 0.2)` |
| `p` | What is the probability of **this or less**? | `pnorm(500, 323, 245)` |
| `q` | Which **value** sits at this probability? | `qnorm(0.95, 323, 245)` |
| `r` | Give me **random** draws from this distribution | `rnorm(100, 323, 245)` |

Attach a prefix to a distribution name: `binom`, `pois`, `norm`, `unif`, `t`.
So `pbinom`, `rpois`, `qnorm` and so on. That is 20 functions from one table.

## Objectives {.smaller}
By the end of this notebook you will be able to:

- Explain what a random variable is, and the difference between discrete and continuous
- Use the binomial distribution to model counts of successes in a fixed number of trials
- Use the Poisson distribution to model counts of events in an interval
- Use the normal distribution to model continuous measurements
- Convert any value to a z-score and interpret it
- Read the `d`/`p`/`q`/`r` function family fluently
- Choose the right distribution for a business situation

## Setup

```{r load}
library(tidyverse)
library(readxl)

sales <- read_excel("../data/supermarket_sales.xlsx")
```

# 1. From describing to modelling

Section 2 described the 1,000 transactions you have. This section does something
different: it builds a **model** of the process that generates transactions, so you
can reason about ones you have not seen yet.

That shift is what makes statistics useful to a business. Describing last month's
sales is bookkeeping. Saying "there is a 5 percent chance we run out of stock next
Tuesday" is planning.

A **random variable** is a numerical outcome of a process whose result is not known
in advance. Two kinds:

| Type | What it can be | Examples here |
|---|---|---|
| **Discrete** | Separate, countable values | `Quantity` (1, 2, 3...), number of complaints |
| **Continuous** | Any value in a range | `Total`, `Unit price`, time between customers |

The distinction matters because the two kinds need different mathematics. For
discrete variables you can ask "what is the probability of exactly 5?" For
continuous variables that question has no useful answer - the probability of a
transaction being exactly 323.0000... is zero - so you ask "what is the probability
of between 300 and 350?" instead.

::: {.callout-tip title="A probability distribution"}
A probability distribution is a rule that assigns probability to each possible
outcome of a random variable.

Think of it as the shape a histogram would settle into if you could collect
infinitely many observations.
:::

# 2. The binomial distribution

Use the binomial when you count **how many successes in a fixed number of
independent attempts**, each with the same probability of success.

Four conditions must hold:

1. A **fixed number** of trials, called `n`
2. Each trial has only **two outcomes** - success or failure
3. The probability of success, `p`, is the **same every time**
4. Trials are **independent** - one outcome does not affect another

## A business question

About half of the chain's customers are members. If 20 customers walk in, what is the
chance that exactly 12 of them are members?

```{r binom_exact}
# First, the actual member rate in our data
p_member <- mean(sales$`Customer type` == "Member")
p_member
```

That line is worth reading carefully. `sales$\`Customer type\` == "Member"` produces
a column of `TRUE`/`FALSE`. R treats `TRUE` as 1 and `FALSE` as 0, so taking the mean
of that column gives the **proportion** that are `TRUE`. This is a very common R
idiom and you will use it constantly.

```{r binom_exact2}
# Probability of exactly 12 members out of 20 customers
dbinom(12, size = 20, prob = p_member)
```

About a 12 percent chance. Now some more useful questions:

```{r binom_cumulative}
# 12 or fewer members
pbinom(12, size = 20, prob = p_member)

# More than 12 members (note: 1 minus "12 or fewer")
1 - pbinom(12, size = 20, prob = p_member)
```

::: {.callout-warning title="The off-by-one trap"}
`pbinom(12, ...)` means "12 **or fewer**", including 12.

So "more than 12" is `1 - pbinom(12, ...)`, **not** `1 - pbinom(13, ...)`. Getting
this wrong by one is the most common error with discrete distributions. When in
doubt, write out the values you want to include.
:::

## Seeing the whole distribution

```{r binom_plot}
binom_dist <- tibble(
  members     = 0:20,
  probability = dbinom(0:20, size = 20, prob = p_member)
)

ggplot(binom_dist, aes(x = members, y = probability)) +
  geom_col(fill = "#A83296") +
  labs(title = "Members among 20 random customers",
       subtitle = paste0("Binomial, n = 20, p = ", round(p_member, 3)),
       x = "Number of members", y = "Probability") +
  theme_minimal()
```

The distribution peaks at 10, which is what you would expect when half of all
customers are members. Its **mean is n times p**, and its spread is
`sqrt(n * p * (1 - p))`:

```{r binom_moments}
n <- 20
tibble(
  mean = n * p_member,
  sd   = sqrt(n * p_member * (1 - p_member))
)
```

:::: {.callout-note title="Your turn"}
About one third of transactions are paid by Ewallet. Out of 15 customers, what is the
probability that **fewer than 4** pay by Ewallet?

```{r ex_binom, eval=FALSE}
p_ewallet <- mean(sales$Payment == "______")

# "Fewer than 4" means 3 or fewer
pbinom(____, size = ____, prob = p_ewallet)
```

::: {.callout-note collapse="true" title="Show the answer"}
```r
p_ewallet <- mean(sales$Payment == "Ewallet")
pbinom(3, size = 15, prob = p_ewallet)
```

The key step is translating "fewer than 4" into "3 or fewer". `pbinom` always
includes the value you give it, so you must subtract one from a strict inequality.
:::
::::

# 3. The Poisson distribution

Use the Poisson when you count **how many events occur in a fixed interval** of time
or space, and there is no natural upper limit.

The binomial asks "how many of these 20?" The Poisson asks "how many will happen at
all?"

| Situation | Distribution |
|---|---|
| Of 20 customers, how many are members? | Binomial (there is a maximum of 20) |
| How many customers arrive in the next hour? | Poisson (no fixed maximum) |
| Of 50 items shipped, how many are damaged? | Binomial |
| How many complaints arrive this week? | Poisson |

The Poisson needs only one number: **lambda**, the average rate of events per
interval.

## A business question

The three branches together handled 1,000 transactions over about 90 days. Suppose a
branch averages 4 transactions per hour. Staffing needs to cope with busy hours.

```{r pois_basic}
lambda <- 4

# Exactly 6 transactions in an hour
dpois(6, lambda = lambda)

# 6 or fewer
ppois(6, lambda = lambda)

# More than 8 - a genuinely busy hour
1 - ppois(8, lambda = lambda)
```

```{r pois_plot}
pois_dist <- tibble(
  transactions = 0:15,
  probability  = dpois(0:15, lambda = lambda)
)

ggplot(pois_dist, aes(x = transactions, y = probability)) +
  geom_col(fill = "#A83296") +
  labs(title = "Transactions per hour",
       subtitle = "Poisson, lambda = 4",
       x = "Transactions in one hour", y = "Probability") +
  theme_minimal()
```

Notice the shape is **right-skewed**. There is a hard floor at zero - you cannot have
negative transactions - but no ceiling, so the tail runs to the right. This is why
count data so often fails a normality check, as you saw in section 2.

::: {.callout-tip title="A useful oddity of the Poisson"}
For a Poisson distribution the **mean and the variance are both lambda**.

That gives you a free diagnostic. If you have count data whose variance is much
larger than its mean, a Poisson model is probably wrong - the events are not
occurring independently, perhaps because they arrive in clusters.
:::

## Why the shape matters for staffing

```{r pois_staffing}
tibble(
  capacity          = 4:10,
  prob_overwhelmed  = round(1 - ppois(4:10, lambda = 4), 4)
)
```

Read that table as a manager. If you staff for the average of 4 transactions an
hour, you are overwhelmed about 37 percent of the time. Staffing for 8 drops that to
around 2 percent.

This is the entire point of probability distributions in business: **planning for the
average guarantees you fail about half the time.** You have to plan for the tail.

::: {.content-visible when-format="revealjs"}
::: {.notes}
Ask the room to guess the overflow probability at capacity 4 before revealing
the table. Most say 10 or 20 percent. The real figure of roughly 37 percent is
the moment the idea lands.
:::
:::

:::: {.callout-note title="Your turn"}
A branch receives an average of 2.5 customer complaints per week. What is the
probability of a week with **no complaints at all**?

```{r ex_pois, eval=FALSE}
dpois(____, lambda = ____)
```

::: {.callout-note collapse="true" title="Show the answer"}
```r
dpois(0, lambda = 2.5)
```

About 8 percent. A complaint-free week is uncommon but not remarkable - which is
worth knowing before anyone declares a quality initiative a success on one good
week's evidence.
:::
::::

# 4. The normal distribution

The most important continuous distribution. It is symmetric, bell-shaped, and
defined by exactly two numbers: the **mean** (where the centre sits) and the
**standard deviation** (how wide it is).

```{r norm_shapes}
norm_demo <- tibble(
  x = seq(-6, 6, length.out = 400)
) |>
  mutate(
    `SD = 1`   = dnorm(x, mean = 0, sd = 1),
    `SD = 2`   = dnorm(x, mean = 0, sd = 2),
    `SD = 0.5` = dnorm(x, mean = 0, sd = 0.5)
  ) |>
  pivot_longer(-x, names_to = "spread", values_to = "density")

ggplot(norm_demo, aes(x = x, y = density, colour = spread)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = c("#A83296", "#D070D0", "#5C1A54")) +
  labs(title = "Three normal distributions with the same mean",
       x = "Value", y = "Density") +
  theme_minimal()
```

All three are centred at zero. The standard deviation alone controls how spread out
they are.

## The empirical rule

For any normal distribution:

| Within | Contains about |
|---|---|
| 1 standard deviation of the mean | 68 percent of values |
| 2 standard deviations | 95 percent |
| 3 standard deviations | 99.7 percent |

Verify it in R rather than taking it on trust:

```{r norm_empirical}
tibble(
  within_sd = 1:3,
  proportion = round(pnorm(1:3) - pnorm(-(1:3)), 4)
)
```

`pnorm(1) - pnorm(-1)` is "the area below +1 minus the area below -1", which is the
area between them. That is the 68 percent.

## Applying it to the ratings data

`Rating` is not perfectly normal - you saw that in section 2 - but it is symmetric
enough to illustrate the method.

```{r norm_rating}
r_mean <- mean(sales$Rating)
r_sd   <- sd(sales$Rating)

tibble(mean = round(r_mean, 3), sd = round(r_sd, 3))
```

```{r norm_rating_q}
# What proportion of customers should rate us below 5, if ratings were normal?
pnorm(5, mean = r_mean, sd = r_sd)

# What proportion actually did?
mean(sales$Rating < 5)
```

The model's prediction and the reality differ. That gap is the model being imperfect,
and noticing it is more valuable than the calculation itself. A model is a
simplification you adopt deliberately, not a truth you discover.

## Going the other way with qnorm

`pnorm` takes a value and returns a probability. `qnorm` does the reverse.

```{r norm_qnorm}
# Which rating do the bottom 10 percent of customers fall below?
qnorm(0.10, mean = r_mean, sd = r_sd)

# Which rating marks the top 5 percent?
qnorm(0.95, mean = r_mean, sd = r_sd)
```

This is how you set thresholds. "Flag any customer rating in the bottom 10 percent
for follow-up" becomes an actual number you can put in a report.

# 5. Z-scores and the standard normal

A **z-score** answers: how many standard deviations is this value from the mean?

$$z = \frac{x - \mu}{\sigma}$$

It converts any normal distribution into the **standard normal** - mean 0, standard
deviation 1 - which is what the printed z-tables in your course folder tabulate.

```{r zscore_manual}
# A customer who left a rating of 9. How unusual is that?
z <- (9 - r_mean) / r_sd
z
```

A z-score of about 1.2 means this rating sits 1.2 standard deviations above average.

```{r zscore_prob}
# What proportion of customers rate us 9 or lower?
pnorm(z)

# So what proportion rate us above 9?
1 - pnorm(z)
```

::: {.callout-tip title="Why z-scores exist"}
They make **incomparable things comparable**.

A student scoring 78 on a maths exam and 84 on a history exam has done better in
history - unless the maths exam was much harder. Convert both to z-scores and you can
say which performance was more impressive *relative to everyone else who sat the same
paper*.

The same logic lets you compare a branch's sales performance with its customer
satisfaction performance, even though one is measured in currency and the other on a
1-to-10 scale.
:::

## Standardising a whole column

```{r zscore_column}
sales_z <- sales |>
  mutate(
    total_z  = (Total - mean(Total)) / sd(Total),
    rating_z = (Rating - mean(Rating)) / sd(Rating)
  )

sales_z |>
  select(`Invoice ID`, Total, total_z, Rating, rating_z) |>
  arrange(desc(total_z)) |>
  head(5)
```

R has a built-in shortcut, `scale()`, though it returns a matrix rather than a plain
column, so the explicit version above is clearer while you are learning.

```{r zscore_check}
# A standardised column always has mean 0 and sd 1
tibble(
  mean_z = round(mean(sales_z$total_z), 10),
  sd_z   = sd(sales_z$total_z)
)
```

::: {.callout-warning title="Standardising does not make data normal"}
Converting to z-scores **shifts and rescales** a distribution. It does not change its
shape.

`Total` was right-skewed in section 2, and `total_z` is exactly as right-skewed.
Students frequently assume standardising fixes non-normality. It does not, and no
amount of z-scoring will make a skewed variable bell-shaped.
:::

## Using z-scores to find unusual transactions

```{r zscore_outliers}
sales_z |>
  filter(abs(total_z) > 2.5) |>
  select(`Invoice ID`, Branch, `Product line`, Total, total_z) |>
  arrange(desc(total_z))
```

`abs()` takes the absolute value, so this catches unusually large *and* unusually
small values in one condition.

This is a second outlier rule to sit alongside the 1.5 x IQR rule from section 2.
They disagree sometimes, and neither is authoritative - the z-score rule is pulled
around by the very outliers it is trying to find, because extreme values inflate the
standard deviation used to compute it.

# 6. Choosing the right distribution

This is the part that matters in practice, and the part exams test.

::: {.content-visible when-format="revealjs"}
::: {.notes}
Do not give them a list of examples to memorise. Give them the one question -
is there a fixed number of trials? - and make them apply it. The list runs out;
the question does not.
:::
:::

| If you are counting... | And... | Use |
|---|---|---|
| Successes in a fixed number of trials | Each trial is independent, same probability | **Binomial** |
| Events in an interval of time or space | No fixed maximum, constant average rate | **Poisson** |
| A continuous measurement | Symmetric, most values near the centre | **Normal** |
| Any outcome equally likely in a range | No value preferred | **Uniform** |

Some worked judgements:

- *Of 200 emails sent, how many are opened?* Fixed number, two outcomes, roughly constant probability. **Binomial.**
- *How many customers enter the shop between 2pm and 3pm?* Counting events in an interval, no maximum. **Poisson.**
- *What will tomorrow's total revenue be?* A continuous amount built from many independent transactions. **Normal** is a reasonable first model.
- *How many defective units in a batch of 500?* Fixed number of trials. **Binomial.**
- *How many machine breakdowns this month?* Events in an interval. **Poisson.**

:::: {.callout-note title="Self-check"}
A call centre wants to model how many calls arrive between 9am and 10am. A colleague
suggests using a binomial distribution "because each minute either has a call or does
not". What is wrong with that reasoning?

::: {.callout-note collapse="true" title="Show the answer"}
Two things.

First, **a minute can contain more than one call.** The binomial requires each trial
to be a clean success or failure, but "this minute had 3 calls" breaks that.

Second, and more fundamentally, **there is no fixed number of trials.** You could
slice the hour into minutes, or seconds, or milliseconds, and the answer should not
depend on how finely you chop up time. That the model changes when you change an
arbitrary choice is a sign the model is wrong.

The Poisson distribution is precisely the limit of the binomial as the trials get
infinitely small and numerous, which is why it is the right tool for events in
continuous time.
:::
::::

# 7. Simulation - checking your reasoning

The `r` functions generate random draws. This lets you verify an answer you worked
out with formulas, which is the single best habit you can build in this course.

```{r sim_binom}
set.seed(2024)

# Simulate 10,000 groups of 20 customers each
simulated <- rbinom(10000, size = 20, prob = p_member)

# How often were exactly 12 members?
mean(simulated == 12)

# Compare with the exact formula
dbinom(12, size = 20, prob = p_member)
```

The two agree to about two decimal places. If they had not, one of them would be
wrong, and you would know to look again.

```{r sim_plot}
tibble(members = simulated) |>
  ggplot(aes(x = members)) +
  geom_bar(fill = "#A83296") +
  labs(title = "10,000 simulated groups of 20 customers",
       x = "Number of members in the group", y = "Frequency") +
  theme_minimal()
```

::: {.callout-tip title="set.seed()"}
Random numbers in R are **reproducibly** random. `set.seed(2024)` fixes the starting
point, so you and your classmates get identical results from the same code.

Always set a seed in work you will hand in. An analysis nobody can reproduce - not
even you, tomorrow - is not an analysis.
:::

# Your turn - full exercises {.smaller}
**Exercise 1 - Binomial.**
Historically 30 percent of customers buy a product from the Health and beauty line.
Out of the next 25 customers:

a. What is the probability that exactly 8 buy from that line?
b. What is the probability that 10 or more do?
c. What is the expected number, and the standard deviation?

```{r full_ex1, eval=FALSE}
# YOUR CODE HERE
```

**Exercise 2 - Poisson.**
A branch averages 3 refund requests per day.

a. What is the probability of a day with exactly 5 refund requests?
b. What is the probability of a day with more than 6?
c. Produce a table showing, for daily capacities of 3 through 9, the probability of being overwhelmed. What capacity would you recommend and why?

```{r full_ex2, eval=FALSE}
# YOUR CODE HERE
```

**Exercise 3 - Normal.**
Using the actual mean and standard deviation of `Total`:

a. What proportion of transactions should exceed 800, if totals were normal?
b. What proportion actually do?
c. Explain the gap. Which section 2 finding predicts it?

```{r full_ex3, eval=FALSE}
# YOUR CODE HERE
```

**Exercise 4 - Z-scores.**
Find the transaction with the highest `Total`. Compute its z-score and state in one
sentence how unusual it is. Then compute the z-score of the highest `Rating` and say
which of the two is the more extreme observation.

```{r full_ex4, eval=FALSE}
# YOUR CODE HERE
```

**Exercise 5 - Choose and justify.**
For each situation, name the distribution and justify it in one sentence:

a. Number of items in a customer's basket
b. Number of customers who abandon a queue in an hour
c. Total revenue on a randomly chosen day
d. Number of the next 40 online orders that are returned

```{r full_ex5, eval=FALSE}
# YOUR CODE HERE - and write your reasoning as comments
```

Solutions are in `3.1-Probability-Distributions-SOLUTIONS.Rmd`.

# Reference card {.smaller}
| Code | What it does |
|---|---|
| `dbinom(k, n, p)` | P(exactly k successes) |
| `pbinom(k, n, p)` | P(k or fewer successes) |
| `1 - pbinom(k, n, p)` | P(more than k successes) |
| `rbinom(N, n, p)` | N random binomial draws |
| `dpois(k, lambda)` | P(exactly k events) |
| `ppois(k, lambda)` | P(k or fewer events) |
| `rpois(N, lambda)` | N random Poisson draws |
| `dnorm(x, mean, sd)` | Height of the normal curve at x |
| `pnorm(x, mean, sd)` | P(value is x or less) |
| `qnorm(p, mean, sd)` | The value below which p of the data falls |
| `rnorm(N, mean, sd)` | N random normal draws |
| `(x - mean(x)) / sd(x)` | Convert to z-scores |
| `mean(vec == "A")` | Proportion of a column equal to "A" |
| `set.seed(2024)` | Make random results reproducible |

# Summary {.smaller}
- A **random variable** is a numerical outcome of an uncertain process; it is either **discrete** or **continuous**.
- The **binomial** counts successes in a fixed number of independent trials.
- The **Poisson** counts events in an interval with no fixed maximum; its mean equals its variance.
- The **normal** models continuous measurements and needs only a mean and a standard deviation.
- The **empirical rule** - 68, 95, 99.7 percent within 1, 2, 3 standard deviations - is worth memorising.
- A **z-score** expresses a value as a distance in standard deviations, making unlike things comparable. It does not change the shape of the distribution.
- R's `d`/`p`/`q`/`r` prefixes work identically for every distribution.
- **Simulate to check your formulas.** If the two disagree, one of them is wrong.

**Next:** section 3.2, where you use these distributions to put a margin of error
around an estimate.

::: {.content-visible when-format="revealjs"}
# Questions?

Work through the exercises in the notebook version of this section before
the next lecture.

::: {.notes}
Point them at the notebook rather than the slides for homework - the slides
drop the longer explanations deliberately.
:::
:::
