---
title: "3.2 Estimation and Confidence Intervals"
subtitle: "307102 Descriptive Statistics for Business | University of Petra"

# Two outputs from this one file: the notebook page, and the lecture deck.
#
# Content that should differ between them is wrapped in a div:
#   ::: {.content-visible unless-format="revealjs"}   notebook prose
#   ::: {.content-visible when-format="revealjs"}     slide version
#
# NOTE: do NOT use when-format="html" for the notebook half. Revealjs is an
# HTML-derived format, so it matches "html" too and BOTH versions appear on
# the slides. Always pair when-format="revealjs" with unless-format="revealjs".
format:
  html: default
  revealjs:
    output-file: 3.2-Estimation-and-Confidence-Intervals-slides.html
    theme: [default, ../assets/petra-reveal.scss]
    width: 1280
    height: 800
    slide-number: c/t
    show-slide-number: all
    scrollable: true
    chalkboard: true
    preview-links: auto
    history: false
    footer: "307102 · Section 3.2 · Estimation · University of Petra"
---

::: {.content-visible unless-format="revealjs"}
::: {.callout-tip title="Lecture slides"}
This same material is also available as a slide deck:
[**Open the slides →**](3.2-Estimation-and-Confidence-Intervals-slides.html)

Arrow keys to move, **F** for full screen, **S** for the presenter view with
speaker notes, **B** for the chalkboard, **Esc** for a grid of all slides.
:::
:::

## New R this week {.smaller}
| Function | What it does | Why a business needs it |
|---|---|---|
| `sample()` | Draws a random subset of rows or values | Simulating what a survey would have found |
| `replicate()` | Repeats an expression many times | Building a sampling distribution from scratch |
| `qt()` | Critical value from the t-distribution | The multiplier in a confidence interval |
| `t.test()` | Confidence interval for a mean, in one line | The everyday tool |
| `prop.test()` | Confidence interval for a proportion | "What share of customers are members?" |

## Objectives {.smaller}
By the end of this notebook you will be able to:

- Explain the difference between a population and a sample, and between a parameter and a statistic
- Describe what a sampling distribution is, and demonstrate the Central Limit Theorem by simulation
- Calculate and interpret the standard error of the mean
- Build a confidence interval for a mean by hand and with `t.test()`
- Build a confidence interval for a proportion
- Explain what a 95 percent confidence interval does and does not mean
- Work out the sample size needed for a target margin of error

## Setup

```{r load}
library(tidyverse)
library(readxl)

set.seed(2024)

sales <- read_excel("../data/supermarket_sales.xlsx")
```

# 1. Population, sample, parameter, statistic

Four words that get confused constantly. Getting them straight is most of the battle.

| Term | Meaning | Notation | Example |
|---|---|---|---|
| **Population** | Every member of the group you care about | - | All transactions the chain will ever process |
| **Sample** | The subset you actually observed | - | Our 1,000 transactions |
| **Parameter** | A number describing the **population** | Greek: mu, sigma, p | The true average transaction value |
| **Statistic** | A number describing the **sample** | Latin: x-bar, s, p-hat | Our sample mean of 323 |

::: {.callout-tip title="The whole of inferential statistics in one sentence"}
You can calculate a **statistic**. You want to know a **parameter**. Statistics is
the discipline of saying how close the first is likely to be to the second.
:::

For teaching purposes we are going to treat our 1,000 transactions as if they were
the entire population, then draw samples from them. That lets us see the true answer
and check whether our methods find it - something you can never do with real data,
which is exactly why simulation is such a good teacher.

```{r population}
true_mean <- mean(sales$Total)
true_sd   <- sd(sales$Total)

tibble(true_mean = round(true_mean, 2), true_sd = round(true_sd, 2))
```

# 2. Sampling variation

Take a sample of 50 transactions and compute the mean. Then do it again.

```{r sampling_variation}
sample_1 <- sample(sales$Total, size = 50)
sample_2 <- sample(sales$Total, size = 50)
sample_3 <- sample(sales$Total, size = 50)

tibble(
  sample = c("Sample 1", "Sample 2", "Sample 3", "TRUE POPULATION"),
  mean   = round(c(mean(sample_1), mean(sample_2), mean(sample_3), true_mean), 2)
)
```

Three different samples, three different answers, none of them exactly the true
value. **This is not a mistake. It is the fundamental problem of statistics.**

Your business has one sample. It got one of these numbers. The question is not "how
do I avoid this?" - you cannot - but "how far off am I likely to be?"

# 3. The sampling distribution

If one sample mean is unreliable, what happens if we look at thousands of them?

```{r sampling_dist}
sample_means_50 <- replicate(2000, mean(sample(sales$Total, size = 50)))

tibble(
  mean_of_the_means = round(mean(sample_means_50), 2),
  true_mean         = round(true_mean, 2),
  sd_of_the_means   = round(sd(sample_means_50), 2)
)
```

Two remarkable things.

**The average of the sample means lands almost exactly on the true mean.** Individual
samples are unreliable, but they are unreliable *symmetrically* - they miss high as
often as low. The sample mean is an **unbiased** estimator.

**The sample means vary far less than individual transactions do.** The transactions
had a standard deviation of about 246; the sample means have one of about 35.
Averaging cancels out noise.

```{r sampling_dist_plot}
tibble(sample_mean = sample_means_50) |>
  ggplot(aes(x = sample_mean)) +
  geom_histogram(bins = 40, fill = "#A83296", colour = "white") +
  geom_vline(xintercept = true_mean, colour = "#1A1A1A", linewidth = 1) +
  labs(title = "2,000 sample means, each from a sample of 50",
       subtitle = "The solid line is the true population mean",
       x = "Sample mean", y = "Frequency") +
  theme_minimal()
```

# 4. The Central Limit Theorem

Look at that histogram again. It is **bell-shaped** - even though individual
transaction totals were clearly right-skewed in section 2.

That is the Central Limit Theorem, and it is the single most important result in
statistics.

::: {.callout-tip title="The Central Limit Theorem"}
As the sample size grows, the distribution of the **sample mean** approaches a normal
distribution - **regardless of the shape of the original population**.

Its centre is the population mean. Its standard deviation is
`population sd / sqrt(n)`, which is called the **standard error**.
:::

This is why the normal distribution dominates statistics even though real business
data is rarely normal. You are almost never modelling individual values. You are
modelling an average, and averages become normal even when the things being averaged
are not.

## Watching it happen

```{r clt_demo}
sample_sizes <- c(2, 5, 30, 100)

clt_data <- map_dfr(sample_sizes, function(size_n) {
  tibble(
    n           = paste("n =", size_n),
    sample_mean = replicate(2000, mean(sample(sales$Total, size = size_n)))
  )
}) |>
  mutate(n = factor(n, levels = paste("n =", sample_sizes)))

ggplot(clt_data, aes(x = sample_mean)) +
  geom_histogram(bins = 40, fill = "#A83296", colour = "white") +
  facet_wrap(~ n, scales = "free") +
  labs(title = "The Central Limit Theorem in action",
       subtitle = "Skewed at n = 2, unmistakably normal by n = 30",
       x = "Sample mean", y = "Frequency") +
  theme_minimal()
```

At `n = 2` the distribution still carries the right skew of the original data. By
`n = 30` it is convincingly bell-shaped. This is where the common rule of thumb
"n of at least 30" comes from.

::: {.callout-warning title="n = 30 is a rule of thumb, not a law"}
How large a sample you need depends on **how skewed the population is**. Mildly
skewed data is fine at n = 15. Severely skewed data - insurance claims, say, where
one claim can dwarf a thousand others - may need several hundred.

The honest procedure is to check, not to recite 30.
:::

## The standard error

```{r standard_error}
n <- 50

tibble(
  formula_se  = round(true_sd / sqrt(n), 3),
  simulated_se = round(sd(sample_means_50), 3)
)
```

The formula and the simulation agree. The **standard error** is the standard
deviation of the sampling distribution: the typical distance between a sample mean
and the truth.

Notice the `sqrt(n)`. To halve your uncertainty you must **quadruple** your sample
size. This is the single most important practical fact about sampling, and it is why
surveys get expensive fast.

::: {.content-visible when-format="revealjs"}
::: {.notes}
The practical version: a survey ten times bigger is only about three times more
precise. Someone will one day ask you to approve that budget, and this is the
slide that tells you what you are buying.
:::
:::

```{r se_by_n}
tibble(
  sample_size    = c(10, 25, 50, 100, 400, 1600),
  standard_error = round(true_sd / sqrt(c(10, 25, 50, 100, 400, 1600)), 2)
)
```

Going from 100 to 400 costs four times as much and only halves the error. Going from
400 to 1600 costs four times again for another halving. There is a point where more
data stops being worth the money, and this table is how you find it.

# 5. Confidence intervals for a mean

A single number - "the average transaction is 323" - hides all of this uncertainty. A
**confidence interval** reports a range instead.

## Building one by hand

The recipe: take your estimate, and go out a certain number of standard errors either
side.

$$\bar{x} \pm t^* \times \frac{s}{\sqrt{n}}$$

```{r ci_manual}
my_sample <- sample(sales$Total, size = 50)

x_bar <- mean(my_sample)
s     <- sd(my_sample)
n     <- length(my_sample)
se    <- s / sqrt(n)

# The multiplier for 95 percent confidence, with n - 1 degrees of freedom
t_star <- qt(0.975, df = n - 1)

tibble(
  x_bar  = round(x_bar, 2),
  se     = round(se, 2),
  t_star = round(t_star, 3),
  lower  = round(x_bar - t_star * se, 2),
  upper  = round(x_bar + t_star * se, 2)
)
```

Three things to notice about `qt(0.975, df = n - 1)`.

**Why 0.975 and not 0.95?** For a 95 percent interval you leave 5 percent outside,
split as 2.5 percent in each tail. So you want the value with 97.5 percent below it.

**Why the t-distribution rather than the normal?** Because you are using the *sample*
standard deviation `s` in place of the unknown population sigma. That substitution
adds uncertainty, and the t-distribution's slightly fatter tails account for it.

**Why `n - 1`?** These are the degrees of freedom. Once you know the sample mean and
`n - 1` of the values, the last value is determined - so only `n - 1` pieces of
information are genuinely free.

```{r t_vs_normal}
tibble(
  sample_size = c(5, 10, 30, 100, 1000),
  t_multiplier = round(qt(0.975, df = c(5, 10, 30, 100, 1000) - 1), 3),
  z_multiplier = round(qnorm(0.975), 3)
)
```

The t multiplier starts large and converges on 1.96 as the sample grows. With small
samples you pay a real penalty in interval width; by n = 100 the difference is
negligible.

## The one-line version

```{r ci_ttest}
t.test(my_sample)
```

`t.test()` does everything above. Read the `95 percent confidence interval` line.
(The rest of that output is a hypothesis test, which is section 3.3 - ignore it for
now.)

To extract just the interval:

```{r ci_extract}
t.test(my_sample)$conf.int
```

Different confidence levels:

```{r ci_levels}
tibble(
  level = c("90%", "95%", "99%"),
  lower = c(t.test(my_sample, conf.level = 0.90)$conf.int[1],
            t.test(my_sample, conf.level = 0.95)$conf.int[1],
            t.test(my_sample, conf.level = 0.99)$conf.int[1]),
  upper = c(t.test(my_sample, conf.level = 0.90)$conf.int[2],
            t.test(my_sample, conf.level = 0.95)$conf.int[2],
            t.test(my_sample, conf.level = 0.99)$conf.int[2])
) |>
  mutate(width = round(upper - lower, 2),
         lower = round(lower, 2),
         upper = round(upper, 2))
```

**Higher confidence means a wider interval.** There is no free lunch: the only way to
be more certain of capturing the truth is to claim less about where it is. A 100
percent confidence interval would run from minus infinity to plus infinity - perfectly
reliable and perfectly useless.

# 6. What a confidence interval actually means

This is the most misunderstood idea in the course, and it is worth getting exactly
right.

::: {.callout-warning title="What it does NOT mean"}
A 95 percent confidence interval does **not** mean "there is a 95 percent probability
the true mean is in this interval".

The true mean is a fixed number. It is either in your interval or it is not. There is
no probability about it once the interval is calculated.
:::

::: {.callout-tip title="What it DOES mean"}
The **procedure** captures the true mean 95 percent of the time.

If you repeated the whole exercise - draw a sample, build an interval - a hundred
times, about 95 of your intervals would contain the true value and about 5 would
miss.

You do not know which kind yours is.
:::

::: {.content-visible when-format="revealjs"}
::: {.notes}
This is the single most examinable idea in the section and the one students get
wrong most often. Run the coverage plot live and have them count the red
intervals. Saying it is not enough - they have to see the method fail.
:::
:::

## Demonstrating it

```{r ci_coverage}
set.seed(99)

ci_results <- map_dfr(1:100, function(i) {
  s  <- sample(sales$Total, size = 50)
  ci <- t.test(s)$conf.int
  tibble(
    run      = i,
    lower    = ci[1],
    upper    = ci[2],
    captured = ci[1] <= true_mean & true_mean <= ci[2]
  )
})

# How many of the 100 intervals contained the true mean?
sum(ci_results$captured)
```

```{r ci_coverage_plot}
ggplot(ci_results, aes(y = run, colour = captured)) +
  geom_segment(aes(x = lower, xend = upper, yend = run), linewidth = 0.5) +
  geom_vline(xintercept = true_mean, colour = "#1A1A1A", linewidth = 0.9) +
  scale_colour_manual(values = c("TRUE" = "#A83296", "FALSE" = "#D64545"),
                      name = "Captured the truth") +
  labs(title = "100 confidence intervals from 100 different samples",
       subtitle = "The vertical line is the true population mean",
       x = "Transaction total", y = "Sample number") +
  theme_minimal()
```

Each horizontal line is one 95 percent confidence interval from one sample. Most
cross the true mean; a handful, shown in red, miss entirely.

**Those red intervals are not mistakes.** Nothing was done wrong in producing them.
They are the 5 percent the method is designed to get wrong, and there is no way to
tell from the inside which of your intervals is one of them.

# 7. Confidence intervals for a proportion

Not everything is a mean. Often you want a **share**: what percentage of customers
are members?

```{r ci_prop}
members <- sum(sales$`Customer type` == "Member")
total   <- nrow(sales)

tibble(members, total, proportion = round(members / total, 4))
```

```{r ci_prop_test}
prop.test(members, total)
```

`prop.test(successes, trials)` gives the interval directly.

```{r ci_prop_extract}
prop.test(members, total)$conf.int
```

The interval comfortably includes 0.50, so on this evidence you could not rule out
that the chain's customer base is exactly half members - a useful thing to know
before anyone builds a strategy on a supposed imbalance.

## The margin of error

```{r ci_prop_moe}
p_hat <- members / total

moe <- 1.96 * sqrt(p_hat * (1 - p_hat) / total)

tibble(
  estimate       = round(p_hat, 4),
  margin_of_error = round(moe, 4),
  lower          = round(p_hat - moe, 4),
  upper          = round(p_hat + moe, 4)
)
```

This is the "plus or minus 3 percentage points" you see quoted with opinion polls.
Now you know exactly where that number comes from.

# 8. How big a sample do I need?

Turn the formula around. If you want a margin of error of `E`:

$$n = \left(\frac{1.96 \times \sigma}{E}\right)^2$$

```{r sample_size_mean}
target_errors <- c(50, 25, 10, 5)

tibble(
  target_margin = target_errors,
  needed_n      = ceiling((1.96 * true_sd / target_errors)^2)
)
```

To estimate average transaction value to within 5 either way, you need a sample of
several thousand. To within 50, a few dozen. **The cost of precision rises as the
square of the precision you demand.**

For a proportion, the worst case is `p = 0.5`, which makes planning easy:

```{r sample_size_prop}
target_moe <- c(0.05, 0.03, 0.02, 0.01)

tibble(
  target_margin = target_moe,
  needed_n      = ceiling(1.96^2 * 0.25 / target_moe^2)
)
```

That second row is why national opinion polls almost always sample around 1,000
people. It is the cheapest sample size that delivers a 3-point margin of error - and
notice that it does not depend on the size of the population at all. Polling Jordan
and polling China need the same sample size for the same precision, which surprises
almost everyone.

:::: {.callout-note title="Self-check"}
A colleague says: "our survey of 1,000 customers has a 3 percent margin of error, so
if we surveyed 10,000 the margin would drop to 0.3 percent." What is wrong?

::: {.callout-note collapse="true" title="Show the answer"}
The margin of error shrinks with the **square root** of the sample size, not with the
sample size itself.

Multiplying `n` by 10 divides the margin by `sqrt(10)`, which is about 3.16. So the
margin would fall from about 3 percent to about 0.95 percent - a real improvement, but
nothing like the tenfold reduction the colleague expects.

To actually reach 0.3 percent you would need roughly 100 times the original sample -
about 100,000 people.

This misunderstanding costs real money. Someone who believes error falls linearly
with sample size will happily approve a survey ten times larger expecting ten times
the precision, and will be disappointed.
:::
::::

# Your turn - full exercises {.smaller}
**Exercise 1 - Sampling distribution.**
Draw 1,000 samples of size 25 from `Rating` and store their means. Report the mean and
standard deviation of those sample means, compare the standard deviation with the
theoretical standard error, and plot the distribution.

```{r full_ex1, eval=FALSE}
# YOUR CODE HERE
```

**Exercise 2 - Confidence interval by hand and by function.**
Take a sample of 80 transactions. Build a 95 percent confidence interval for the mean
`Total` **by hand** using `qt()`, then confirm it with `t.test()`. Do the two agree?

```{r full_ex2, eval=FALSE}
# YOUR CODE HERE
```

**Exercise 3 - Confidence level and width.**
Using the same sample of 80, build 80, 90, 95 and 99 percent intervals. Produce a
table of their widths and write one sentence explaining the pattern.

```{r full_ex3, eval=FALSE}
# YOUR CODE HERE
```

**Exercise 4 - Proportion.**
What proportion of transactions are paid by credit card? Build a 95 percent confidence
interval. Could the true proportion plausibly be one third?

```{r full_ex4, eval=FALSE}
# YOUR CODE HERE
```

**Exercise 5 - Sample size planning.**
Your manager wants to estimate the average customer `Rating` to within 0.1 either way,
with 95 percent confidence. Using the sample standard deviation of `Rating` as an
estimate of sigma, how many customers must be surveyed? If each survey costs 2 dinars,
what is the budget - and what would it be for a margin of 0.2 instead?

```{r full_ex5, eval=FALSE}
# YOUR CODE HERE
```

Solutions are in `3.2-Estimation-and-Confidence-Intervals-SOLUTIONS.Rmd`.

# Reference card {.smaller}
| Code | What it does |
|---|---|
| `sample(x, size = n)` | Draw a random sample |
| `sample(x, size = n, replace = TRUE)` | Draw with replacement (bootstrap) |
| `replicate(N, expr)` | Repeat an expression N times |
| `sd(x) / sqrt(n)` | Standard error of the mean |
| `qt(0.975, df = n - 1)` | t multiplier for a 95 percent interval |
| `qnorm(0.975)` | z multiplier, 1.96 |
| `t.test(x)` | Confidence interval for a mean |
| `t.test(x, conf.level = 0.99)` | A different confidence level |
| `t.test(x)$conf.int` | Just the interval |
| `prop.test(successes, trials)` | Confidence interval for a proportion |
| `ceiling((1.96 * sd / E)^2)` | Sample size for margin of error E |

# Summary {.smaller}
- A **statistic** describes your sample; a **parameter** describes the population. You compute the first to estimate the second.
- Sample means vary from sample to sample. That variation is the **sampling distribution**.
- The **Central Limit Theorem** says the sampling distribution of the mean becomes normal as `n` grows, whatever the shape of the population.
- The **standard error**, `s / sqrt(n)`, is the typical distance from a sample mean to the truth. Halving it requires **quadrupling** the sample.
- A **confidence interval** reports a range instead of a single number. Use `t.test()` for means, `prop.test()` for proportions.
- A 95 percent interval means **the procedure** succeeds 95 percent of the time. It does not assign a probability to your particular interval.
- Higher confidence buys width, not accuracy. Required sample size grows with the **square** of the precision you demand.

**Next:** section 3.3, where you stop estimating a value and start testing a claim
about it.

::: {.content-visible when-format="revealjs"}
# Questions?

Work through the exercises in the notebook version of this section before
the next lecture.

::: {.notes}
Point them at the notebook rather than the slides for homework - the slides
drop the longer explanations deliberately.
:::
:::
