---
title: "3.2 Estimation and Confidence Intervals - Solutions"
subtitle: "307102 Descriptive Statistics for Business | University of Petra"
code-fold: true
code-summary: "Show the code"
---

::: {.callout-caution title="Instructor copy"}
Every chunk runs. Results involving `sample()` vary between runs unless a seed is
set; quoted figures come from `set.seed(2024)`.
:::

```{r load}
library(tidyverse)
library(readxl)

set.seed(2024)

sales <- read_excel("../data/supermarket_sales.xlsx")

true_mean <- mean(sales$Total)
true_sd   <- sd(sales$Total)
```

# Inline exercises

## Section 8 - The square root misunderstanding

Covered in the student notebook. The point to land: **margin of error scales with
`1/sqrt(n)`**, so tenfold more data gives about a 3.16-fold improvement, not tenfold.

Worth putting on the board:

```{r sol_sqrt}
tibble(
  n              = c(1000, 4000, 10000, 100000),
  margin_percent = round(1.96 * sqrt(0.25 / c(1000, 4000, 10000, 100000)) * 100, 2)
)
```

**Why this matters commercially.** Someone who believes error falls linearly with
sample size will approve a survey ten times larger expecting ten times the precision.
That is a real budget decision made on a false premise, and it happens often.

# Exercise 1 - Sampling distribution

```{r sol_ex1}
sample_means <- replicate(1000, mean(sample(sales$Rating, size = 25)))

tibble(
  mean_of_means   = round(mean(sample_means), 4),
  true_mean       = round(mean(sales$Rating), 4),
  sd_of_means     = round(sd(sample_means), 4),
  theoretical_se  = round(sd(sales$Rating) / sqrt(25), 4)
)
```

```{r sol_ex1_plot}
tibble(sample_mean = sample_means) |>
  ggplot(aes(x = sample_mean)) +
  geom_histogram(bins = 40, fill = "#A83296", colour = "white") +
  geom_vline(xintercept = mean(sales$Rating), colour = "#1A1A1A", linewidth = 1) +
  labs(title = "1,000 sample means of Rating, each from a sample of 25",
       x = "Sample mean rating", y = "Frequency") +
  theme_minimal()
```

**Expected results.** The mean of the sample means lands within about 0.01 of the true
mean of 6.97. The simulated standard deviation is close to the theoretical standard
error of about 0.34.

**The teaching point students miss.** `Rating` was **not** normal in section 2 - it was
flat and bounded, with strongly negative excess kurtosis. Yet the distribution of its
sample means is convincingly bell-shaped.

That is the Central Limit Theorem doing exactly what it promises, on a population that
looks nothing like a bell curve. Ask the class to compare this histogram with the
histogram of raw `Rating` from section 2. The contrast makes the theorem concrete in a
way the formal statement never does.

**Marking guidance.** Full credit needs the comparison between the simulated standard
deviation and the theoretical standard error. Students who produce the histogram
without checking that the two agree have done the mechanics without the verification,
which is the habit this course is trying to build.

# Exercise 2 - Confidence interval by hand and by function

```{r sol_ex2}
set.seed(2024)
my_sample <- sample(sales$Total, size = 80)

x_bar  <- mean(my_sample)
s      <- sd(my_sample)
n      <- length(my_sample)
se     <- s / sqrt(n)
t_star <- qt(0.975, df = n - 1)

by_hand <- c(x_bar - t_star * se, x_bar + t_star * se)

by_function <- as.numeric(t.test(my_sample)$conf.int)

tibble(
  method = c("By hand", "t.test()"),
  lower  = round(c(by_hand[1], by_function[1]), 4),
  upper  = round(c(by_hand[2], by_function[2]), 4)
)
```

**They agree to the last decimal place**, because `t.test()` computes exactly this
formula.

**Why this exercise exists.** Students who only ever call `t.test()` treat it as
magic, and cannot debug it when the answer looks wrong. Doing it once by hand
converts the function from an oracle into a shortcut.

**Marking guidance.** Common errors, in order of frequency:

1. Using `qt(0.95, ...)` instead of `qt(0.975, ...)` - forgetting the interval is
   two-tailed. This is the most common by a wide margin.
2. Using `df = n` instead of `df = n - 1`.
3. Dividing by `n` instead of `sqrt(n)` in the standard error.
4. Using `qnorm()` instead of `qt()`. At n = 80 this barely changes the answer, which
   makes it a good moment to ask *when* it would matter - and the answer is small
   samples.

# Exercise 3 - Confidence level and width

```{r sol_ex3}
levels <- c(0.80, 0.90, 0.95, 0.99)

map_dfr(levels, function(lv) {
  ci <- t.test(my_sample, conf.level = lv)$conf.int
  tibble(
    level = paste0(lv * 100, "%"),
    lower = round(ci[1], 2),
    upper = round(ci[2], 2),
    width = round(ci[2] - ci[1], 2)
  )
})
```

**Expected pattern.** Width increases monotonically with confidence level, and the
increase **accelerates** - the step from 95 to 99 percent costs far more width than
the step from 80 to 90.

**Model sentence.**

> Higher confidence produces a wider interval, because the only way to be more certain
> of capturing the true mean is to make a less specific claim about where it is. The
> cost accelerates: going from 95 to 99 percent confidence widens the interval by more
> than going from 80 to 90 does, because you are buying coverage from the increasingly
> thin tails of the distribution.

**Discussion prompt.** Ask what a 100 percent confidence interval would look like. The
answer - minus infinity to plus infinity - makes the trade-off unforgettable. Perfect
reliability and zero information are the same thing.

# Exercise 4 - Proportion

```{r sol_ex4}
credit <- sum(sales$Payment == "Credit card")
total  <- nrow(sales)

tibble(credit, total, proportion = round(credit / total, 4))

prop.test(credit, total)$conf.int
```

**Expected results.** 311 of 1,000 transactions, a proportion of 0.311, with a 95
percent interval of roughly 0.283 to 0.341.

**The answer to the question:** one third is 0.3333, which **falls inside** the
interval. So yes - the true proportion could plausibly be exactly one third. The data
does not rule it out.

**Marking guidance.** The interpretation is the mark. Look for the student explicitly
checking whether 0.3333 lies between the bounds, and stating the conclusion in terms
of plausibility rather than proof.

Watch for the reverse error: a student concluding the proportion **is** one third.
The interval is consistent with one third, but equally consistent with 0.29 or 0.34.
Failing to rule something out is not the same as establishing it - a distinction that
becomes central in the next section.

# Exercise 5 - Sample size planning

```{r sol_ex5}
sigma <- sd(sales$Rating)
sigma

n_for <- function(E) ceiling((1.96 * sigma / E)^2)

tibble(
  margin_of_error = c(0.05, 0.1, 0.2, 0.5),
  needed_n        = n_for(c(0.05, 0.1, 0.2, 0.5)),
  cost_dinars     = n_for(c(0.05, 0.1, 0.2, 0.5)) * 2
)
```

**Expected results.** With `sigma` of about 1.719, a margin of 0.1 needs **1,135**
respondents costing **2,270 dinars**. A margin of 0.2 needs **284** respondents
costing **568 dinars**.

**The finding worth stating plainly:** **doubling the acceptable margin cuts the cost
to a quarter.**

**Marking guidance.** The arithmetic is easy; the recommendation is the exercise. A
strong answer challenges the specification rather than merely costing it:

> A margin of 0.1 on a 1-to-10 rating scale costs about 2,270 dinars. Relaxing the
> target to 0.2 costs about 568 - a saving of roughly 1,700 dinars. Since the observed
> differences in average rating between branches and product lines are themselves only
> around 0.2 to 0.3, a margin of 0.1 is finer than any decision we would actually make
> on this data. I would recommend the 0.2 target and spend the difference elsewhere.

That answer connects the statistical requirement to the business decision, which is
the whole point of sample size planning. A student who computes 1,140 and stops has
done the sum but not the thinking.

::: {.callout-caution title="A caveat worth mentioning"}
This calculation uses the sample standard deviation as if it were the population
sigma, which is circular - you would not know it before running the survey.

In practice you take sigma from a pilot study, from historical data, or from a
published figure for a similar measure. Mentioning this stops the sharper students
from thinking they have spotted a flaw, and introduces the idea of a pilot study at no
extra cost.
:::

# Teaching notes

**Where students get stuck.**

1. **The interpretation of a confidence interval.** Expect the "95 percent probability
   the mean is in here" phrasing repeatedly, including from students who can recite the
   correct version. The coverage plot in section 6 is the most effective remedy - run
   it live and let them count the red intervals.
2. **`qt(0.975)` versus `qt(0.95)`.** Draw the two tails on the board every time.
3. **Confusing the standard deviation with the standard error.** The one-line
   distinction that works: *the standard deviation describes the data; the standard
   error describes the estimate.*
4. **Believing a bigger population needs a bigger sample.** The polling example in
   section 8 is the fastest cure. Sample size depends on the variability of the thing
   measured and the precision you want, not on how many people exist.

**Timing.** Sections 1 to 4 fill one 90-minute session and are the conceptual core.
Sections 5 to 8 fill a second and are more procedural. If pressed, section 8 can be
set as reading, but do not cut section 6 - the interpretation of a confidence interval
is the single most examinable idea in the notebook.

**A demonstration worth the five minutes.** Before showing any theory, have the class
each draw a sample of 20 from the data and write their sample mean on the board. The
spread of thirty different answers to the same question makes the sampling
distribution real in a way no simulation quite matches, because the numbers are
theirs.

**Connecting to the Excel track.** `CONFIDENCE.T(alpha, sd, n)` returns the margin of
error, which you then add and subtract manually; `t.test()` returns the finished
interval. `T.INV.2T(0.05, df)` is `qt(0.975, df)`. Excel has no direct equivalent of
`prop.test()`, which is one of the clearer wins for R in this section.
