---
title: "3.3 Hypothesis Testing - Solutions"
subtitle: "307102 Descriptive Statistics for Business | University of Petra"
code-fold: true
code-summary: "Show the code"
---

::: {.callout-caution title="Instructor copy"}
Every chunk runs.

**A warning about this notebook before you teach it.** Almost every test in these
exercises comes back **non-significant**. That is not a defect in the dataset - it is
the most valuable feature of it.

Students arrive expecting statistics to produce findings. What it mostly produces is
the disciplined conclusion that an apparent pattern is indistinguishable from noise.
A course where every test is significant teaches students to expect significance, and
they then go and find it whether it is there or not.

Say this out loud in the first session. Otherwise students conclude they have done
something wrong.
:::

```{r load}
library(tidyverse)
library(readxl)
library(car)
library(effectsize)

sales <- read_excel("../data/supermarket_sales.xlsx")
```

# Inline exercises

## Section 8 - The manager running 20 tests

```{r sol_multiple}
tibble(
  n_tests = c(1, 5, 10, 20, 50),
  prob_at_least_one_false_positive = round(1 - 0.95^c(1, 5, 10, 20, 50), 3)
)
```

**The point.** With 20 tests there was about a 64 percent chance of at least one
"significant" result even if every branch were identical. A single p = 0.04 out of 20
tests is exactly what pure noise looks like.

**What he should do.** Either correct for the number of tests:

```{r sol_padjust}
# Twenty p-values, one of which is 0.04
example_pvals <- c(0.04, runif(19, 0.06, 0.99))

round(min(p.adjust(example_pvals, method = "holm")), 3)
```

After correction the smallest adjusted p-value is nowhere near significance.

Or - better - treat the finding as a **hypothesis to test on fresh data**. A result
that survives on new data is a finding; one that does not was noise. This is the
distinction between exploratory and confirmatory analysis, and it is worth naming
explicitly.

# Exercise 1 - One-sample t-test

**Hypotheses.**

- H0: the population mean rating equals 7.0
- H1: the population mean rating does not equal 7.0

```{r sol_ex1}
t.test(sales$Rating, mu = 7.0)
```

```{r sol_ex1_summary}
tibble(
  sample_mean = round(mean(sales$Rating), 4),
  p_value     = round(t.test(sales$Rating, mu = 7)$p.value, 4)
)
```

**Expected result.** The sample mean is about 6.97, and the p-value is large - nowhere
near 0.05. **Fail to reject H0.**

**Model conclusion.**

> Average customer satisfaction is 6.97 out of 10. This is not significantly different
> from management's claimed figure of 7.0, so the data provides no evidence against it.

**Marking guidance.** Three things to look for.

The student must write **"fail to reject"**, not **"accept"**. We have not shown the
mean is 7.0; we have shown the data is consistent with it, which is weaker.

The confidence interval should be mentioned. It contains 7.0, which is the same
conclusion viewed from a different angle - and reinforces that the test and the
interval are one calculation, not two.

The conclusion must be in **business language**. "t = -0.50, p = 0.617" is not a
conclusion. A manager cannot act on it.

::: {.callout-caution title="Discussion prompt"}
Ask: *would we have rejected if management had claimed 7.2?*

Running it takes ten seconds and the answer is yes. That makes the point that a
hypothesis test evaluates **a specific claim**, not the data in general - and that
"not significant" is always relative to what was proposed.
:::

# Exercise 2 - Two-sample with assumptions and effect size

```{r sol_ex2_explore}
sales |>
  group_by(Gender) |>
  summarise(
    n          = n(),
    mean_spend = round(mean(Total), 2),
    sd_spend   = round(sd(Total), 2)
  )
```

```{r sol_ex2_plot}
ggplot(sales, aes(x = Gender, y = Total)) +
  geom_boxplot(fill = "#A83296", alpha = 0.7) +
  labs(title = "Transaction value by gender", x = NULL, y = "Transaction total") +
  theme_minimal()
```

```{r sol_ex2_assumptions}
leveneTest(Total ~ factor(Gender), data = sales)
```

```{r sol_ex2_qq}
ggplot(sales, aes(sample = Total)) +
  stat_qq(colour = "#A83296", alpha = 0.4) +
  stat_qq_line(colour = "#1A1A1A") +
  facet_wrap(~ Gender) +
  labs(title = "Q-Q plots by gender", x = "Theoretical", y = "Observed") +
  theme_minimal()
```

```{r sol_ex2_test}
t.test(Total ~ Gender, data = sales)
```

```{r sol_ex2_effect}
cohens_d(Total ~ Gender, data = sales)
```

**Expected results.** Female customers average **335.10** against **310.79** for male
customers, on 501 and 499 transactions. Levene's test gives p = 0.34. The t-test gives
**p = 0.118**, and **Cohen's d is 0.099**.

**The model answer.**

> Female customers spend about 335 per transaction against about 311 for male
> customers. Levene's test finds no evidence of unequal variances, and while the Q-Q
> plots show the right skew we identified in section 2, the sample of roughly 500 per
> group makes the t-test robust to it via the Central Limit Theorem.
>
> The difference is not statistically significant (p = 0.118), and Cohen's d of 0.099
> is well below even the conventional "small" threshold of 0.2. Even if it were
> significant, a gap of this size explains almost none of the variation in what
> customers spend, and would not justify a gender-differentiated marketing strategy.

**Marking guidance.** Full credit requires the student to report the effect size **and
use it**. A student who reports d = 0.1 and then recommends a strategy has computed
the number without understanding it.

The strongest answers note that the conclusion would be the same either way: the effect
is too small to act on regardless of which side of 0.05 the p-value lands. That is
exactly the reasoning this section is trying to instil.

# Exercise 3 - ANOVA

```{r sol_ex3_explore}
sales |>
  group_by(Branch) |>
  summarise(
    n         = n(),
    mean_total = round(mean(Total), 2),
    sd_total   = round(sd(Total), 2)
  )
```

```{r sol_ex3_anova}
branch_model <- aov(Total ~ Branch, data = sales)

summary(branch_model)
```

```{r sol_ex3_effect}
eta_squared(branch_model)
```

**Expected results.** Branch means run roughly 312, 320 and 337. The ANOVA p-value is
above 0.05, and eta-squared is tiny - on the order of 0.001 to 0.002.

**Conclusion: fail to reject.** There is no evidence the three branches differ in
average transaction value.

**The interpretation that earns the marks.** Eta-squared of about 0.002 means branch
explains roughly **0.2 percent** of the variation in transaction value. Even if the
test had been significant, 99.8 percent of why transactions differ has nothing to do
with which branch they occurred in.

That is a far more useful sentence for a manager than "p = 0.23".

**On TukeyHSD.** The exercise says to run it *if the result is significant*. It is
not, so a student who runs it anyway has missed a step in the logic - though running it
to see that all intervals contain zero is a reasonable way to confirm the conclusion.
Accept it with a note.

```{r sol_ex3_tukey}
# Shown for completeness - every interval contains zero
TukeyHSD(branch_model)
```

::: {.callout-caution title="Worth connecting back"}
In section 2, Exercise 1, students computed these same branch means and were asked
whether the differences mattered. The strong answers said no, informally, by comparing
the gap between branches with the spread inside them.

This ANOVA is the formal version of that intuition. Making the connection explicit is
worth two minutes - it shows the course is one argument rather than a sequence of
techniques.
:::

# Exercise 4 - Chi-square

```{r sol_ex4_table}
ct <- table(sales$`Customer type`, sales$`Product line`)

ct
```

```{r sol_ex4_expected}
result <- chisq.test(ct)

# Assumption check: all expected counts should be at least 5
round(result$expected, 1)

min(result$expected)
```

```{r sol_ex4_test}
result
```

```{r sol_ex4_residuals}
round(result$residuals, 2)
```

**Expected results.** All expected counts comfortably exceed 5, so the assumption
holds. The p-value is above 0.05, and no residual approaches the rough threshold of
plus or minus 2.

**Conclusion: fail to reject.** There is no evidence that members and non-members buy
from different product lines.

**Why the residual step matters even when the test is non-significant.** It confirms
that no single cell is driving anything, which is a stronger statement than the
p-value alone. Had one residual been 3.5 with an overall p of 0.06, that would be worth
investigating on new data.

**Marking guidance.** The assumption check is not optional and should carry marks.
Students who run `chisq.test()` without examining `$expected` have skipped the one step
that determines whether the p-value means anything.

**A commercially useful reading.** "Membership does not predict what people buy" is a
genuine finding with an operational consequence: the loyalty programme should not be
used to target product-specific promotions, because members are not a distinct
purchasing segment. Encourage students to state the *business* implication of a null
result rather than treating it as a non-answer.

# Exercise 5 - Choose and defend

| Question | Test | Justification |
|---|---|---|
| a. Morning versus afternoon ratings | **Independent two-sample t-test** | Numeric outcome, two independent groups of different customers. If time were split into three or more periods, ANOVA instead. |
| b. Proportion of members across three branches | **Chi-square test of independence** | Two categorical variables - branch and customer type - and counts, not means. |
| c. Rating before and after training, same 40 customers | **Paired t-test** | Each customer is measured twice, so the two samples are not independent. |
| d. Gross income across six product lines | **One-way ANOVA** | Numeric outcome across more than two groups. |

**Marking guidance.** The justification carries the marks, and it must name the
*structural* feature: how many groups, whether the outcome is numeric or categorical,
and whether the observations are paired.

**Part (c) is the discriminator.** A student who says "two-sample t-test" has missed
the pairing. This is worth dwelling on because it is the error with real consequences -
analysing paired data as independent throws away statistical power and can turn a real
effect into a null result. The demonstration in section 3 of the student notebook shows
exactly how large that loss can be.

**Part (b) catches a different error.** Some students propose a t-test because "we are
comparing proportions and proportions are numbers". The outcome is *membership*, which
is categorical; the proportion is a summary of it, not the raw measurement. A
`prop.test()` comparing two branches would be defensible, but with three branches
chi-square is the natural choice.

# Teaching notes

**Where students get stuck.**

1. **Interpreting the p-value.** Expect "the probability the null is true" from
   students who can recite the correct definition. The phrasing that helps: *a p-value
   assumes the null is true - it cannot also tell you whether it is.*
2. **"Accept the null."** Correct this every single time, from week one. The habit
   forms early or not at all.
3. **Believing non-significant means no effect.** It means insufficient evidence, which
   may simply reflect a small sample. Distinguishing "no effect" from "no evidence of
   an effect" is a genuinely difficult idea and needs repetition.
4. **Skipping assumption checks** because the test runs without complaining. R will
   happily return a p-value from a badly violated test.
5. **The formula notation `y ~ x`.** Unfamiliar at first, but it recurs in ANOVA and
   regression, so time spent here pays off twice.

**Timing.** Sections 1 to 5 fill one 90-minute session - the logic, t-tests,
assumptions and effect size. Sections 6 to 8 fill a second - ANOVA, chi-square and test
selection. Section 5 is the one not to cut.

**The demonstration to run live.** The effect size table in section 5, where a trivial
difference becomes significant as `n` grows. Run it in front of the class and let them
watch the p-value fall while the effect size does not move. Nothing else makes the
point as economically.

**Connecting to the Excel track.** `T.TEST(range1, range2, tails, type)` covers the
t-tests, with `type` 1, 2 and 3 for paired, equal-variance and unequal-variance. Excel's
Analysis ToolPak has single-factor ANOVA. `CHISQ.TEST(observed, expected)` requires you
to build the expected table yourself, which `chisq.test()` does automatically.

Excel has **no built-in effect size**, no Levene's test, no Shapiro-Wilk, and no
post-hoc correction. That gap is the clearest argument in the whole course for teaching
this material in R.
