---
title: "3.3 Hypothesis Testing"
subtitle: "307102 Descriptive Statistics for Business | University of Petra"

# Two outputs from this one file: the notebook page, and the lecture deck.
#
# Content that should differ between them is wrapped in a div:
#   ::: {.content-visible unless-format="revealjs"}   notebook prose
#   ::: {.content-visible when-format="revealjs"}     slide version
#
# NOTE: do NOT use when-format="html" for the notebook half. Revealjs is an
# HTML-derived format, so it matches "html" too and BOTH versions appear on
# the slides. Always pair when-format="revealjs" with unless-format="revealjs".
format:
  html: default
  revealjs:
    output-file: 3.3-Hypothesis-Testing-slides.html
    theme: [default, ../assets/petra-reveal.scss]
    width: 1280
    height: 800
    slide-number: c/t
    show-slide-number: all
    scrollable: true
    chalkboard: true
    preview-links: auto
    history: false
    footer: "307102 · Section 3.3 · Hypothesis Testing · University of Petra"
---

::: {.content-visible unless-format="revealjs"}
::: {.callout-tip title="Lecture slides"}
This same material is also available as a slide deck:
[**Open the slides →**](3.3-Hypothesis-Testing-slides.html)

Arrow keys to move, **F** for full screen, **S** for the presenter view with
speaker notes, **B** for the chalkboard, **Esc** for a grid of all slides.
:::
:::

## New R this week {.smaller}
| Function | What it does | When you reach for it |
|---|---|---|
| `t.test()` | Compares a mean to a value, or two means | "Do members spend more than non-members?" |
| `aov()` | Compares three or more means | "Do the six product lines differ in rating?" |
| `TukeyHSD()` | Says *which* groups differ, after `aov()` | "Fine - but which two?" |
| `chisq.test()` | Tests association between two categorical variables | "Do branches differ in payment method?" |
| `shapiro.test()`, `leveneTest()` | Check the assumptions before you trust the test | Always |
| `cohens_d()`, `eta_squared()` | Effect size - is the difference big enough to care? | Always |

## Objectives {.smaller}
By the end of this notebook you will be able to:

- State a null and alternative hypothesis for a business question
- Explain what a p-value is, and what it is not
- Run and interpret one-sample, paired and independent t-tests
- Run a one-way ANOVA and follow it with a post-hoc test
- Run a chi-square test of independence
- Check the assumptions each test requires
- Report an effect size alongside every p-value, and explain why that matters

## Setup

```{r load}
library(tidyverse)
library(readxl)
library(car)
library(effectsize)

sales <- read_excel("../data/supermarket_sales.xlsx")
```

# 1. The logic of a hypothesis test

Section 3.2 estimated values. This section tests **claims**.

The reasoning is deliberately backwards, and that is what makes it hard at first.

::: {.callout-tip title="The argument, in four steps"}
1. Assume **nothing is going on**. This is the null hypothesis.
2. Work out what data you would expect to see if that were true.
3. Compare it with the data you actually got.
4. If your data would be very surprising under that assumption, **reject** it.
:::

It is proof by contradiction with a probability attached. You never prove the
alternative; you only show the null is an uncomfortable place to stand.

::: {.content-visible when-format="revealjs"}
::: {.notes}
Warn them explicitly that the logic runs backwards. Students expect a test to
prove the interesting thing; it only ever makes the boring thing uncomfortable
to keep believing.
:::
:::

| Hypothesis | Symbol | What it says | Example |
|---|---|---|---|
| **Null** | H0 | No effect, no difference, no association | Members and non-members spend the same |
| **Alternative** | H1 | There is an effect | They do not spend the same |

Two rules students break constantly:

- The null hypothesis always contains the **equals** sign.
- You test the null, but you **never accept** it. You either reject it or you fail to reject it, which is not the same as proving it true.

## The p-value

The p-value is the probability of getting data **at least as extreme as yours**, *if
the null hypothesis were true*.

::: {.callout-warning title="What a p-value is NOT"}
It is **not** the probability the null hypothesis is true.

It is **not** the probability your result happened by chance.

It is **not** a measure of how large or important the effect is.

A p-value is a statement about *data given a hypothesis*, never about *a hypothesis
given data*. Reversing those two is the most common error in applied statistics, and
it appears in published research constantly.
:::

By convention, `p < 0.05` counts as evidence against the null. That threshold, called
alpha, is arbitrary - it was a suggestion by Ronald Fisher in the 1920s, not a law of
nature. A p-value of 0.049 and one of 0.051 mean essentially the same thing.

## Two ways to be wrong

| | H0 is actually true | H0 is actually false |
|---|---|---|
| **You reject H0** | **Type I error** (false alarm) | Correct |
| **You fail to reject H0** | Correct | **Type II error** (missed it) |

Setting alpha to 0.05 means accepting a 5 percent Type I error rate. Lowering it to
0.01 reduces false alarms but increases missed real effects. You cannot minimise both
at once, so the right threshold depends on which mistake is more costly in context -
a question statistics cannot answer for you.

# 2. One-sample t-test

**The question:** management claims the average transaction is 300. Our sample says
otherwise. Is the difference real?

- H0: the population mean equals 300
- H1: the population mean does not equal 300

```{r one_sample}
t.test(sales$Total, mu = 300)
```

Reading that output:

| Line | Meaning |
|---|---|
| `t = ...` | The test statistic: how many standard errors the sample mean sits from 300 |
| `df = 999` | Degrees of freedom, `n - 1` |
| `p-value` | Probability of a result this extreme if the true mean really were 300 |
| `95 percent confidence interval` | The plausible range for the true mean |
| `mean of x` | Our sample mean |

**The conclusion in business language.** The interval does not contain 300 and the
p-value is small, so the average transaction value is higher than management's figure.
Note that the confidence interval and the test always agree: if the interval excludes
the null value, the test rejects it. They are two views of the same calculation.

## One-tailed tests

If you only care about one direction, say so:

```{r one_tailed}
t.test(sales$Total, mu = 300, alternative = "greater")
```

::: {.callout-warning title="Decide the direction before you look"}
A one-tailed test has more power to detect an effect in the direction you specified -
but you must choose that direction **before** seeing the data.

Running a two-tailed test, noticing the result points one way, and then switching to
one-tailed to halve the p-value is a form of cheating with a name: **p-hacking**. It
roughly doubles your false positive rate.
:::

# 3. Two-sample t-test

**The question:** do members spend more than non-members?

- H0: the two group means are equal
- H1: they are not

```{r two_sample_explore}
sales |>
  group_by(`Customer type`) |>
  summarise(
    n         = n(),
    mean_spend = round(mean(Total), 2),
    sd_spend   = round(sd(Total), 2)
  )
```

Always look before you test.

```{r two_sample_plot}
ggplot(sales, aes(x = `Customer type`, y = Total)) +
  geom_boxplot(fill = "#A83296", alpha = 0.7) +
  labs(title = "Transaction value by customer type",
       x = NULL, y = "Transaction total") +
  theme_minimal()
```

```{r two_sample_test}
t.test(Total ~ `Customer type`, data = sales)
```

The formula `Total ~ \`Customer type\`` reads as "Total **broken down by** customer
type". You will see this `~` notation constantly from here on.

R runs **Welch's t-test** by default, which does not assume the two groups have equal
variances. This is the right default and you should almost always keep it.

## Paired t-test

When the two measurements come from the **same** unit - before and after, left and
right, same customer twice - the samples are not independent and you must say so.

```{r paired_demo}
set.seed(2024)

# 30 customers rated us before and after a service change
before <- rnorm(30, mean = 6.8, sd = 1.2)
after  <- before + rnorm(30, mean = 0.4, sd = 0.8)

t.test(before, after, paired = TRUE)
```

```{r paired_wrong}
# The same data analysed WRONGLY as if independent
t.test(before, after, paired = FALSE)
```

Compare the two p-values. The paired test is far more sensitive, because it removes
the variation *between customers* and looks only at the change *within* each one.

::: {.callout-tip title="The pairing question"}
Ask: **could I shuffle one group without losing information?**

If each value in group A is meaningfully tied to a specific value in group B, the test
is paired. Getting this wrong throws away real statistical power - or invents power
you do not have.
:::

# 4. Checking assumptions

Every test has conditions. Reporting a p-value from a test whose assumptions fail is
reporting a number that does not mean what you say it means.

The t-test assumes:

1. Observations are **independent** - a design question, not a statistical one
2. The data is roughly **normal** within each group, or the sample is large enough for the CLT
3. For the classic pooled test, **equal variances** - but Welch's test removes this

## Normality

```{r assump_normality}
member_spend <- sales |> filter(`Customer type` == "Member") |> pull(Total)

shapiro.test(member_spend)
```

Shapiro-Wilk tests H0: the data is normal. A small p-value means **reject normality**.

::: {.callout-warning title="The paradox of normality tests"}
With a large sample, a normality test rejects almost everything. Real data is never
exactly normal, and with 500 observations even trivial departures become
"statistically significant".

So a significant Shapiro-Wilk on a large sample tells you very little. **Look at a Q-Q
plot instead** and ask whether the departure is big enough to matter.
:::

```{r assump_qq}
ggplot(sales, aes(sample = Total)) +
  stat_qq(colour = "#A83296", alpha = 0.4) +
  stat_qq_line(colour = "#1A1A1A") +
  facet_wrap(~ `Customer type`) +
  labs(title = "Q-Q plots of transaction total by customer type",
       x = "Theoretical quantiles", y = "Observed") +
  theme_minimal()
```

Both groups curve upward at the right - the right skew you found in section 2. With
roughly 500 observations per group, the Central Limit Theorem makes the t-test robust
to this. The skew would be a genuine problem at n = 15.

## Equal variances

```{r assump_levene}
leveneTest(Total ~ factor(`Customer type`), data = sales)
```

Levene's test has H0: the variances are equal. A large p-value means no evidence
against equality.

Since we are using Welch's test, which does not assume equal variances anyway, this is
informational rather than decisive.

# 5. Effect size - the question p-values cannot answer

A p-value tells you whether an effect exists. It says nothing about whether it
**matters**.

::: {.callout-tip title="Why this section exists"}
With a large enough sample, **any** difference becomes statistically significant -
including differences far too small to act on.

A test can tell you a branch's average sale is genuinely 2 dinars higher. Only you can
decide 2 dinars is not worth a strategy.
:::

```{r effect_size}
cohens_d(Total ~ `Customer type`, data = sales)
```

Cohen's d expresses the difference between two means in standard deviations - the same
idea as a z-score, applied to a gap between groups.

| Cohen's d | Conventional label |
|---|---|
| 0.2 | Small |
| 0.5 | Medium |
| 0.8 | Large |

These labels are conventions, not laws. In a context where a tiny effect is
enormously valuable - a 0.1 percent improvement in conversion on millions of visits -
a "small" effect can be worth a great deal.

## The sample size demonstration

```{r effect_demo}
set.seed(2024)

# A deliberately trivial difference: 0.02 of a standard deviation
demo_test <- function(n) {
  a <- rnorm(n, mean = 100, sd = 15)
  b <- rnorm(n, mean = 100.3, sd = 15)
  tibble(
    n       = n,
    p_value = round(t.test(a, b)$p.value, 4),
    cohens_d = round(as.numeric(cohens_d(a, b)$Cohens_d), 4)
  )
}

map_dfr(c(50, 500, 5000, 50000), demo_test)
```

The effect size stays trivially small at every sample size. The p-value marches
steadily towards zero.

**This is the single most important table in the notebook.** It shows that
"statistically significant" and "practically important" are different questions, and
that only one of them is answered by a p-value.

::: {.content-visible when-format="revealjs"}
::: {.notes}
Run this live. Watching the p-value march toward zero while Cohen's d does not
move is worth more than any amount of explaining. Then ask which column they
would put in a report.
:::
:::

Always report both.

# 6. ANOVA - comparing three or more groups

**The question:** do the six product lines differ in customer rating?

You might think to run all 15 pairwise t-tests. Do not. Each test has a 5 percent
false positive rate, so across 15 tests the chance of at least one false alarm is
about 54 percent.

```{r anova_multiple}
# Probability of at least one false positive across 15 independent tests
round(1 - 0.95^15, 3)
```

ANOVA asks the question once: **are all the group means equal?**

```{r anova_explore}
sales |>
  group_by(`Product line`) |>
  summarise(
    n           = n(),
    mean_rating = round(mean(Rating), 3),
    sd_rating   = round(sd(Rating), 3)
  ) |>
  arrange(desc(mean_rating))
```

```{r anova_plot}
ggplot(sales, aes(x = `Product line`, y = Rating)) +
  geom_boxplot(fill = "#A83296", alpha = 0.7) +
  coord_flip() +
  labs(title = "Customer rating by product line", x = NULL, y = "Rating") +
  theme_minimal()
```

::: {.callout-warning title="Give the column a name without a space first"}
Backticks work fine in `aov()` itself, but some of the functions you run *afterwards* -
`TukeyHSD()` in particular - cannot cope with a column name containing a space, and
fail with the unhelpful message `undefined columns selected`.

The fix takes one line: make a copy of the column with a simple name before you model
it. Do this whenever you are about to fit a model on a column whose name has a space.
:::

```{r anova_prep}
sales_anova <- sales |>
  mutate(product_line = factor(`Product line`))
```

```{r anova_test}
anova_model <- aov(Rating ~ product_line, data = sales_anova)

summary(anova_model)
```

Reading the ANOVA table:

| Column | Meaning |
|---|---|
| `Df` | Degrees of freedom: groups minus 1, then the remainder |
| `Sum Sq` | Variation between groups, then within groups |
| `Mean Sq` | Sum of squares divided by degrees of freedom |
| `F value` | Between-group variation divided by within-group variation |
| `Pr(>F)` | The p-value |

The intuition behind F: if the groups genuinely differ, the variation *between* group
means should be large relative to the variation *within* groups. F near 1 means the
groups are no more different from each other than random members of the same group.

## Effect size for ANOVA

```{r anova_effect}
eta_squared(anova_model)
```

Eta-squared is the **proportion of variation in the outcome explained by the grouping**.
Read a value of 0.01 as "product line explains 1 percent of the variation in ratings",
which leaves 99 percent explained by something else.

## Post-hoc testing

ANOVA says *whether* the groups differ, not *which* ones. For that you need a post-hoc
test that corrects for multiple comparisons.

```{r anova_tukey}
TukeyHSD(anova_model)
```

Tukey's HSD compares every pair while holding the overall false positive rate at 5
percent. Read the `p adj` column - the adjusted p-value - and the confidence interval
for each difference. **Any interval containing zero means that pair is not
distinguishable.**

## ANOVA assumptions

```{r anova_assumptions}
leveneTest(Rating ~ product_line, data = sales_anova)
```

```{r anova_residuals}
ggplot(tibble(residual = residuals(anova_model)), aes(sample = residual)) +
  stat_qq(colour = "#A83296", alpha = 0.4) +
  stat_qq_line(colour = "#1A1A1A") +
  labs(title = "Q-Q plot of ANOVA residuals",
       x = "Theoretical quantiles", y = "Residual") +
  theme_minimal()
```

ANOVA assumes independent observations, roughly normal residuals, and roughly equal
variances across groups. With group sizes over 150 it is fairly robust to moderate
departures.

# 7. Chi-square test of independence

Everything so far compared **means**. The chi-square test works with **counts**, and
asks whether two categorical variables are associated.

**The question:** do the three branches differ in how customers pay?

- H0: branch and payment method are independent
- H1: they are associated

```{r chisq_table}
payment_table <- table(sales$Branch, sales$Payment)

payment_table
```

```{r chisq_test}
chisq.test(payment_table)
```

## What the test is actually doing

```{r chisq_expected}
chi_result <- chisq.test(payment_table)

# What we would expect if branch and payment were completely independent
round(chi_result$expected, 1)
```

```{r chisq_residuals}
# Where the observed counts deviate most from expectation
round(chi_result$residuals, 2)
```

The residuals show which cells drive the result. Values beyond about +2 or -2 mark
cells where the observed count differs notably from what independence predicts. This
turns a single p-value into an actual finding you can describe.

::: {.callout-warning title="The chi-square assumption"}
Chi-square needs **expected** counts of at least 5 in each cell - check
`chi_result$expected`, not the observed table.

If some cells are too small, either combine categories or use
`fisher.test()` instead.
:::

## Goodness of fit

A different chi-square question: does one categorical variable match an expected
distribution?

```{r chisq_gof}
# Are the three branches equally busy?
chisq.test(table(sales$Branch))
```

By default this tests whether all categories are equally likely.

# 8. Choosing the right test

| Your question | Data type | Test |
|---|---|---|
| Is this mean different from a fixed value? | One numeric variable | `t.test(x, mu = value)` |
| Do two independent groups differ? | Numeric outcome, 2 groups | `t.test(y ~ group)` |
| Did the same units change? | Numeric, measured twice | `t.test(a, b, paired = TRUE)` |
| Do three or more groups differ? | Numeric outcome, 3+ groups | `aov(y ~ group)` |
| Which specific groups differ? | After a significant ANOVA | `TukeyHSD()` |
| Are two categorical variables associated? | Two categorical variables | `chisq.test(table(a, b))` |
| Does one categorical variable match expectations? | One categorical variable | `chisq.test(table(a))` |

:::: {.callout-note title="Self-check"}
A manager runs 20 separate tests looking for anything that differs between branches.
One comes back with p = 0.04. He announces a finding. What is wrong?

::: {.callout-note collapse="true" title="Show the answer"}
With 20 tests at alpha = 0.05, the expected number of false positives **when nothing
is going on at all** is exactly one.

```r
1 - 0.95^20   # about 0.64
```

There was roughly a 64 percent chance of finding at least one "significant" result
even if every branch were identical. Finding exactly one at p = 0.04 is precisely what
pure noise looks like.

This is the **multiple comparisons problem**, and it is why ANOVA exists rather than
running every pairwise t-test.

What should he do? Either correct for the number of tests - `p.adjust()` does this - or
treat the finding as a **hypothesis to test on fresh data** rather than a conclusion.
The second is almost always better science.
:::
::::

# Your turn - full exercises {.smaller}
**Exercise 1 - One-sample.**
Management claims average customer satisfaction is 7.0 out of 10. Test this claim.
State your hypotheses, run the test, and write a one-sentence conclusion in business
language.

```{r full_ex1, eval=FALSE}
# YOUR CODE HERE
```

**Exercise 2 - Two-sample with assumptions and effect size.**
Do male and female customers spend different amounts? Explore first with a table and a
boxplot, check assumptions, run the test, report Cohen's d, and state whether the
result is both statistically significant and practically important.

```{r full_ex2, eval=FALSE}
# YOUR CODE HERE
```

**Exercise 3 - ANOVA.**
Does average transaction value differ across the three branches? Run the ANOVA, report
eta-squared, and if the result is significant, run `TukeyHSD()` to identify which pairs
differ.

```{r full_ex3, eval=FALSE}
# YOUR CODE HERE
```

**Exercise 4 - Chi-square.**
Is there an association between `Customer type` and `Product line`? Build the table,
check the expected counts, run the test, and examine the residuals to say where any
association comes from.

```{r full_ex4, eval=FALSE}
# YOUR CODE HERE
```

**Exercise 5 - Choose and defend.**
For each question, name the test and justify it in one sentence:

a. Do customers rate us differently in the morning than the afternoon?
b. Is the proportion of members the same in all three branches?
c. Did our average rating improve after a training programme, measured on the same 40 customers before and after?
d. Do the six product lines generate different average gross income?

```{r full_ex5, eval=FALSE}
# YOUR CODE HERE - reasoning as comments
```

Solutions are in `3.3-Hypothesis-Testing-SOLUTIONS.Rmd`.

# Reference card {.smaller}
| Code | What it does |
|---|---|
| `t.test(x, mu = 300)` | One-sample t-test |
| `t.test(x, mu = 300, alternative = "greater")` | One-tailed |
| `t.test(y ~ group, data = d)` | Independent two-sample (Welch) |
| `t.test(a, b, paired = TRUE)` | Paired t-test |
| `aov(y ~ group, data = d)` | One-way ANOVA |
| `summary(model)` | The ANOVA table |
| `TukeyHSD(model)` | Post-hoc pairwise comparisons |
| `chisq.test(table(a, b))` | Test of independence |
| `chisq.test(table(a))` | Goodness of fit |
| `result$expected`, `result$residuals` | Diagnose a chi-square result |
| `shapiro.test(x)` | Normality test |
| `leveneTest(y ~ factor(g), data = d)` | Equal variance test |
| `cohens_d(y ~ group, data = d)` | Effect size for two means |
| `eta_squared(aov_model)` | Effect size for ANOVA |
| `p.adjust(pvals, method = "holm")` | Correct for multiple tests |

# Summary {.smaller}
- A hypothesis test assumes **nothing is going on**, then asks how surprising your data would be under that assumption.
- A **p-value** is the probability of data this extreme *if the null were true*. It is not the probability the null is true, and not a measure of importance.
- **Type I** error is a false alarm; **Type II** is a missed effect. Lowering one raises the other.
- Use **t-tests** for one or two means, **ANOVA** for three or more, **chi-square** for counts.
- **Paired** tests apply when the same units are measured twice, and are far more sensitive.
- **Check assumptions** - but on large samples read a Q-Q plot rather than a normality test's p-value.
- **Always report an effect size.** With enough data, everything is significant; only effect size tells you whether it matters.
- Running many tests manufactures false positives. That is what ANOVA and post-hoc corrections exist to prevent.

**Next:** section 3.4, where instead of comparing groups you model a relationship and
use it to predict.

::: {.content-visible when-format="revealjs"}
# Questions?

Work through the exercises in the notebook version of this section before
the next lecture.

::: {.notes}
Point them at the notebook rather than the slides for homework - the slides
drop the longer explanations deliberately.
:::
:::
