---
title: "3.4 Linear Regression"
subtitle: "307102 Descriptive Statistics for Business | University of Petra"

# Two outputs from this one file: the notebook page, and the lecture deck.
#
# Content that should differ between them is wrapped in a div:
#   ::: {.content-visible unless-format="revealjs"}   notebook prose
#   ::: {.content-visible when-format="revealjs"}     slide version
#
# NOTE: do NOT use when-format="html" for the notebook half. Revealjs is an
# HTML-derived format, so it matches "html" too and BOTH versions appear on
# the slides. Always pair when-format="revealjs" with unless-format="revealjs".
format:
  html: default
  revealjs:
    output-file: 3.4-Linear-Regression-slides.html
    theme: [default, ../assets/petra-reveal.scss]
    width: 1280
    height: 800
    slide-number: c/t
    show-slide-number: all
    scrollable: true
    chalkboard: true
    preview-links: auto
    history: false
    footer: "307102 · Section 3.4 · Linear Regression · University of Petra"
---

::: {.content-visible unless-format="revealjs"}
::: {.callout-tip title="Lecture slides"}
This same material is also available as a slide deck:
[**Open the slides →**](3.4-Linear-Regression-slides.html)

Arrow keys to move, **F** for full screen, **S** for the presenter view with
speaker notes, **B** for the chalkboard, **Esc** for a grid of all slides.
:::
:::

## New R this week {.smaller}
| Function | What it does | Why a business needs it |
|---|---|---|
| `lm()` | Fits a linear model | "How much does each extra item add to the bill?" |
| `summary()` | Coefficients, significance and R-squared | Reading the model |
| `predict()` | Predicts the outcome for new data | "What would a basket of 7 items cost?" |
| `augment()` | Adds fitted values and residuals to the data | Diagnostics |
| `plot(model)` | Four diagnostic plots in one call | Checking the model is trustworthy |

## Objectives {.smaller}
By the end of this notebook you will be able to:

- Fit a simple linear regression and interpret its slope and intercept in business terms
- Explain what R-squared measures and what it does not
- Fit a multiple regression and interpret a coefficient while holding others constant
- Check the four assumptions of linear regression using residual plots
- Make predictions with appropriate uncertainty
- Recognise when a regression should not be trusted

## Setup

```{r load}
library(tidyverse)
library(readxl)
library(broom)

sales <- read_excel("../data/supermarket_sales.xlsx")
```

# 1. From comparing to predicting

Section 3.3 compared groups. Regression does something more ambitious: it describes
the **relationship** between variables, in a form you can use to predict.

The model is a straight line:

$$y = \beta_0 + \beta_1 x + \varepsilon$$

| Symbol | Name | Meaning in business terms |
|---|---|---|
| `y` | Outcome, dependent variable | The thing you want to explain or predict |
| `x` | Predictor, independent variable | The thing you think explains it |
| `beta_0` | Intercept | Predicted `y` when `x` is zero |
| `beta_1` | Slope | Change in `y` for a one-unit increase in `x` |
| `epsilon` | Error | Everything the model does not capture |

That last term matters. **No model captures everything**, and a model that appears to
is usually wrong.

# 2. Simple linear regression

**The question:** how much does each additional item add to a customer's bill?

Always plot first.

```{r simple_plot}
ggplot(sales, aes(x = Quantity, y = Total)) +
  geom_point(colour = "#A83296", alpha = 0.35) +
  geom_smooth(method = "lm", colour = "#1A1A1A", se = TRUE) +
  labs(title = "Transaction total against basket size",
       x = "Items purchased", y = "Transaction total") +
  theme_minimal()
```

```{r simple_fit}
model1 <- lm(Total ~ Quantity, data = sales)

summary(model1)
```

## Reading the output

**The coefficients table** is the heart of it.

```{r simple_tidy}
tidy(model1)
```

| Column | Meaning |
|---|---|
| `estimate` | The coefficient itself |
| `std.error` | Its uncertainty |
| `statistic` | Estimate divided by standard error |
| `p.value` | Tests whether the true coefficient is zero |

**Interpreting the slope.** Each additional item is associated with an increase of
about that many currency units in the transaction total. Since `Quantity` and
`Unit price` are independent in this data, that slope is roughly the average unit
price times the tax multiplier - a good sanity check that the model is doing something
sensible.

**Interpreting the intercept.** The predicted total when `Quantity` is zero. Here that
is a transaction with no items, which cannot happen - so the intercept is a
mathematical anchor for the line, not a meaningful business quantity.

::: {.callout-warning title="Intercepts are often meaningless"}
The intercept is only interpretable when `x = 0` is a value that could actually occur
and appears somewhere near your data.

Predicting income at age zero, or revenue with zero customers, is extrapolating far
outside the data. The number is still needed to position the line; it just should not
be quoted as a finding.
:::

## R-squared

```{r simple_glance}
glance(model1) |>
  select(r.squared, adj.r.squared, sigma, p.value, nobs)
```

**R-squared is the proportion of variation in `y` explained by the model.** An
R-squared of 0.50 means the model accounts for half the variability in transaction
totals, leaving half to everything else.

::: {.callout-tip title="R-squared does not mean the model is right"}
A high R-squared means the line fits the points closely. It does **not** mean:

- the relationship is causal
- a straight line was the right shape
- the model will predict well on new data
- the assumptions hold

A curved relationship forced through a straight line can still produce a respectable
R-squared while being systematically wrong at both ends. Always look at the residual
plots, which come in section 5.
:::

**What counts as a good R-squared depends entirely on the field.** In physics, 0.99 is
routine. In predicting human behaviour, 0.30 can be excellent. Judge it against what
is achievable for the question, never against an absolute standard.

::: {.content-visible when-format="revealjs"}
::: {.notes}
Resist giving a threshold. Students want to know what counts as good, and any
number you say will be quoted back at you out of context for the rest of the
semester.
:::
:::

# 3. Multiple regression

One predictor is rarely enough. Adding more asks a sharper question.

**The question:** what drives transaction value - basket size, unit price, or both?

```{r multiple_fit}
model2 <- lm(Total ~ Quantity + `Unit price`, data = sales)

summary(model2)
```

```{r multiple_tidy}
tidy(model2)
```

::: {.callout-tip title="The four words that change everything"}
In a multiple regression, each coefficient is the effect of that predictor
**holding the others constant**.

The `Quantity` coefficient now means: among transactions with the *same unit price*,
each extra item adds this much. That is a genuinely different - and more useful -
question than the one simple regression answered.
:::

```{r multiple_compare}
bind_rows(
  glance(model1) |> mutate(model = "Quantity only"),
  glance(model2) |> mutate(model = "Quantity + Unit price")
) |>
  select(model, r.squared, adj.r.squared, sigma)
```

The second model explains far more. That should not surprise you: a bill is
essentially quantity times price, so giving the model both ingredients lets it
reconstruct nearly all of the total.

::: {.callout-warning title="Use adjusted R-squared to compare models"}
Plain R-squared **never decreases** when you add a predictor, even a useless one.
Adding random noise as a variable will nudge it up.

Adjusted R-squared penalises extra predictors, so it can fall when you add something
that does not earn its place. When comparing models with different numbers of
predictors, that is the number to read.
:::

## Adding a categorical predictor

```{r multiple_categorical}
model3 <- lm(Total ~ Quantity + `Unit price` + Branch, data = sales)

tidy(model3)
```

R automatically converts `Branch` into **dummy variables**. With three branches you get
two coefficients: `BranchB` and `BranchC`. Branch A is missing because it is the
**reference level** - each coefficient is the difference from Branch A, holding
everything else constant.

If those coefficients are small and non-significant, branch adds nothing once you know
what was bought - which is a real finding, and one that would be hard to establish any
other way.

# 4. Making predictions

```{r predict_new}
new_baskets <- tibble(
  Quantity     = c(3, 7, 10),
  `Unit price` = c(25, 50, 80)
)

new_baskets |>
  mutate(predicted_total = round(predict(model2, newdata = new_baskets), 2))
```

## Two kinds of interval

```{r predict_intervals}
# Confidence interval: where is the AVERAGE total for baskets like this?
predict(model2, newdata = new_baskets, interval = "confidence") |> round(2)
```

```{r predict_intervals2}
# Prediction interval: where will an INDIVIDUAL transaction fall?
predict(model2, newdata = new_baskets, interval = "prediction") |> round(2)
```

The prediction interval is much wider, and the reason is worth understanding.

A **confidence interval** captures uncertainty about the *average* - and averages are
stable, so the interval is narrow. A **prediction interval** must also account for the
fact that individual transactions scatter around that average, so it is necessarily
wider.

Confusing the two is a common and costly error. If you promise a client that their
individual order will fall within the confidence interval, you will be wrong far more
often than the stated confidence level suggests.

::: {.callout-warning title="Never extrapolate"}
`Quantity` in this dataset runs from 1 to 10. Asking the model to predict a basket of
50 items produces a number, and R will not warn you.

That number is worthless. The linear relationship was only ever observed inside the
range of the data, and there is no evidence it continues outside it.
:::

# 5. Checking the assumptions

Linear regression assumes four things. All four are checkable, and R gives you the
plots in one call.

| Assumption | What it means | How to check |
|---|---|---|
| **Linearity** | The relationship really is a straight line | Residuals vs Fitted |
| **Independence** | Observations do not influence each other | Study design; ordering plots |
| **Homoscedasticity** | Error spread is constant across the range | Scale-Location |
| **Normality of residuals** | Errors are roughly normally distributed | Q-Q plot of residuals |

```{r diagnostics, fig.width=9, fig.height=7}
par(mfrow = c(2, 2))
plot(model2)
par(mfrow = c(1, 1))
```

## Reading each plot

**Residuals vs Fitted** (top left). You want a formless horizontal band around zero.
A curve means the relationship is not linear and you have fitted the wrong shape.

**Q-Q Residuals** (top right). Points on the diagonal mean the residuals are normal.
This is the assumption regression is least sensitive to, especially with large samples.

**Scale-Location** (bottom left). You want a flat line. An upward fan means the errors
grow with the fitted value - **heteroscedasticity** - which makes the standard errors,
and therefore every p-value in the model, unreliable.

**Residuals vs Leverage** (bottom right). Identifies points with disproportionate
influence. Anything outside the dashed Cook's distance lines is worth investigating -
a single point should not be able to move your conclusions.

## Doing it with ggplot

```{r diagnostics_gg}
aug <- augment(model2)

ggplot(aug, aes(x = .fitted, y = .resid)) +
  geom_point(colour = "#A83296", alpha = 0.35) +
  geom_hline(yintercept = 0, colour = "#1A1A1A", linewidth = 0.8) +
  geom_smooth(se = FALSE, colour = "#D64545") +
  labs(title = "Residuals against fitted values",
       subtitle = "Look for a flat red line and an even band of points",
       x = "Fitted value", y = "Residual") +
  theme_minimal()
```

::: {.callout-tip title="What a residual plot is really asking"}
A residual is what the model got wrong. If the model has captured everything
systematic, what remains should be **pure noise** - no pattern, no trend, no changing
spread.

**Any visible pattern in the residuals is information the model failed to use.**
That is the single most useful sentence in this notebook.
:::

::: {.content-visible when-format="revealjs"}
::: {.notes}
Show a clean residual plot and a badly violated one side by side. The contrast
teaches this faster than any description of what to look for.
:::
:::

# 6. When regression misleads

## Correlation is still not causation

Regression finds associations. Nothing about fitting a line establishes that `x`
causes `y`. A model showing that customers who buy more spend more has not shown that
encouraging purchases *causes* spending - the arithmetic guaranteed the association
before any data was collected.

## Confounding

A variable left out of the model can create or hide a relationship. If expensive
product lines happen to be stocked at the busiest branch, a regression of spend on
branch would attribute a product effect to location. The remedy is to include the
confounder - which is exactly what multiple regression is for.

## Multicollinearity

When two predictors are themselves strongly correlated, the model cannot separate their
effects.

```{r multicollinearity}
# Total and gross income are perfectly proportional - a deliberate example
model_bad <- lm(Total ~ Quantity + `gross income`, data = sales)

summary(model_bad)
```

The R-squared is essentially 1 and the fit looks flawless. It is meaningless.
`gross income` is a fixed 4.7619 percent of `Total`, so this model predicts the total
from a rescaling of itself. **A model that predicts a variable from a transformation of
that same variable is circular**, however good the diagnostics look.

Watch for this in real work: it usually arrives disguised, as a variable derived from
the outcome without anyone noticing.

## Outliers and influence

```{r influence}
aug |>
  slice_max(.cooksd, n = 5) |>
  select(Total, Quantity, `Unit price`, .fitted, .resid, .cooksd)
```

Cook's distance measures how much the fitted line would move if a point were deleted.
High values are not errors, but they are worth understanding - a conclusion resting on
one observation is a fragile conclusion.

# Your turn - full exercises {.smaller}
**Exercise 1 - Simple regression.**
Fit a model predicting `gross income` from `Quantity`. Interpret the slope and the
intercept in business terms, report R-squared, and say whether the model is useful.

```{r full_ex1, eval=FALSE}
# YOUR CODE HERE
```

**Exercise 2 - Multiple regression.**
Fit a model predicting `Total` from `Quantity`, `Unit price` and `Rating`. Which
predictors matter? Interpret the `Rating` coefficient carefully - what does it mean to
hold quantity and unit price constant?

```{r full_ex2, eval=FALSE}
# YOUR CODE HERE
```

**Exercise 3 - Diagnostics.**
Produce the four diagnostic plots for your model from Exercise 2. Comment on each
assumption in one sentence. Is there anything that would make you distrust the model?

```{r full_ex3, eval=FALSE}
# YOUR CODE HERE
```

**Exercise 4 - Prediction.**
Using the Exercise 2 model, predict the total for a basket of 6 items at a unit price
of 45 with a rating of 8. Give both a confidence and a prediction interval, and explain
in one sentence which a shop manager should use and why.

```{r full_ex4, eval=FALSE}
# YOUR CODE HERE
```

**Exercise 5 - Model comparison and judgement.**
Build three models predicting `Total` with increasing numbers of predictors. Compare
them using adjusted R-squared. Then write a short paragraph: which model would you
recommend, and does the best-fitting model tell you anything a manager could act on?

```{r full_ex5, eval=FALSE}
# YOUR CODE HERE
```

Solutions are in `3.4-Linear-Regression-SOLUTIONS.Rmd`.

# Reference card {.smaller}
| Code | What it does |
|---|---|
| `lm(y ~ x, data = d)` | Simple linear regression |
| `lm(y ~ x1 + x2, data = d)` | Multiple regression |
| `lm(y ~ x1 * x2, data = d)` | With an interaction term |
| `summary(model)` | Full output with significance and R-squared |
| `tidy(model)` | Coefficients as a tidy table (`broom`) |
| `glance(model)` | One-row model summary (`broom`) |
| `augment(model)` | Data plus fitted values and residuals (`broom`) |
| `coef(model)` | Just the coefficients |
| `confint(model)` | Confidence intervals for coefficients |
| `predict(model, newdata = d)` | Predict for new data |
| `predict(..., interval = "confidence")` | Interval for the mean response |
| `predict(..., interval = "prediction")` | Interval for a single new observation |
| `plot(model)` | Four diagnostic plots |
| `residuals(model)`, `fitted(model)` | Extract residuals and fitted values |
| `geom_smooth(method = "lm")` | Add a fitted line to a ggplot |

# Summary {.smaller}
- Regression models the **relationship** between an outcome and one or more predictors, letting you predict rather than only compare.
- The **slope** is the change in `y` per one-unit change in `x`. The **intercept** is often not meaningful.
- **R-squared** is the proportion of variation explained. It says nothing about causation, correct model shape, or whether assumptions hold.
- In multiple regression each coefficient holds the other predictors **constant** - a different and sharper question.
- Use **adjusted R-squared** when comparing models with different numbers of predictors.
- **Confidence intervals** are for the average response; **prediction intervals** are for an individual case and are much wider.
- Check all four assumptions with residual plots. **Any pattern in the residuals is information the model failed to use.**
- Beware extrapolation, confounding, multicollinearity, and models that predict a variable from a transformation of itself.

This completes the R track. You now have the full path from describing a dataset to
modelling the process behind it - and, more importantly, a working sense of when to
distrust the answer.

::: {.content-visible when-format="revealjs"}
# Questions?

Work through the exercises in the notebook version of this section before
the next lecture.

::: {.notes}
Point them at the notebook rather than the slides for homework - the slides
drop the longer explanations deliberately.
:::
:::
