---
title: "3.4 Linear Regression - Solutions"
subtitle: "307102 Descriptive Statistics for Business | University of Petra"
code-fold: true
code-summary: "Show the code"
---

::: {.callout-caution title="Instructor copy"}
Every chunk runs.

**A structural warning about this dataset.** `Total` is an exact arithmetic function
of `Quantity` and `Unit price` - specifically `Quantity * Unit price * 1.05`. Any
regression of `Total` on both predictors will therefore have an R-squared very close
to 1.

This is a feature for teaching, not a bug. It gives students a model that fits
essentially perfectly and *still* teaches them almost nothing about the business,
which is the most valuable lesson in the section. Make that point explicitly rather
than letting them believe they have built a triumph.
:::

```{r load}
library(tidyverse)
library(readxl)
library(broom)

sales <- read_excel("../data/supermarket_sales.xlsx")
```

# Exercise 1 - Simple regression

```{r sol_ex1}
m1 <- lm(`gross income` ~ Quantity, data = sales)

summary(m1)
```

```{r sol_ex1_tidy}
tidy(m1)

glance(m1) |> select(r.squared, adj.r.squared, sigma, nobs)
```

```{r sol_ex1_plot}
ggplot(sales, aes(x = Quantity, y = `gross income`)) +
  geom_point(colour = "#A83296", alpha = 0.35) +
  geom_smooth(method = "lm", colour = "#1A1A1A") +
  labs(title = "Gross income against basket size",
       x = "Items purchased", y = "Gross income") +
  theme_minimal()
```

**Expected results.** A slope of **2.83**, strongly significant. An intercept of
**-0.19**, not significantly different from zero (p = 0.73). R-squared of **0.498**.

**Model answer.**

> Each additional item in the basket is associated with about 2.83 more in gross
> income. The intercept of -0.19 is not significantly different from zero and is not
> meaningfully interpretable in any case, since a transaction with no items cannot
> occur - though it is reassuring that it lands near zero, which is where a sensible
> model should put it.
>
> The model explains about 50 percent of the variation in gross income, leaving the
> other half to something else - principally the unit price of what was bought, which
> this model ignores entirely.
>
> The model is useful as a rough guide but should not be relied on for individual
> predictions.

**Marking guidance.** Three things carry the marks:

1. The slope stated **with units and direction**, not just as a number.
2. Explicit recognition that the intercept is not interpretable here.
3. R-squared interpreted as *unexplained* variation as well as explained. Students who
   say "R-squared is 0.25" without saying what that leaves out have described the
   output rather than understood it.

**Discussion prompt.** Ask what the missing 50 percent is. The answer - unit price - is
sitting in the dataset, which motivates Exercise 2 naturally.

# Exercise 2 - Multiple regression

```{r sol_ex2}
m2 <- lm(Total ~ Quantity + `Unit price` + Rating, data = sales)

summary(m2)
```

```{r sol_ex2_tidy}
tidy(m2)

glance(m2) |> select(r.squared, adj.r.squared, sigma)
```

**Expected results.** `Quantity` (58.7) and `Unit price` (5.81) are both
overwhelmingly significant. R-squared is **0.891**.

`Rating` has a coefficient of **-2.85 with p = 0.058** - just above the conventional
threshold, and negative.

::: {.callout-caution title="This result is more useful than a clean null"}
`Rating` at p = 0.058 is the single best teaching moment in the notebook, and it is
worth planning the session around it.

Ask the class: *is rating a significant predictor?* The honest answer is that the
question is badly posed. At alpha = 0.05 you fail to reject; at alpha = 0.10 you
reject. Nothing about the underlying reality changes between those two sentences, which
is precisely why treating 0.05 as a bright line is indefensible.

Then ask the better question: **is the effect big enough to matter?** A coefficient of
-2.85 means that among baskets of the same size and price, a one-point higher rating is
associated with about 2.85 *less* spent. Against a mean transaction of 323 and a
standard deviation of 246, that is roughly one percent of a standard deviation per
rating point. It is negligible whichever side of 0.05 it lands on.

This is the effect-size lesson from section 3.3, arriving unprompted in a real model.
:::

**Model answer.**

> Quantity and unit price both strongly predict the transaction total, which is
> unsurprising since the total is calculated from them. Rating has a small negative
> coefficient of -2.85 with a p-value of 0.058 - marginally above the conventional 0.05
> threshold.
>
> That coefficient means that **among transactions with the same basket size and the
> same unit price**, a one-point higher customer rating is associated with about 2.85
> less spent. Given that transaction totals have a standard deviation of about 246,
> this effect is far too small to be of any practical interest, regardless of where its
> p-value sits relative to 0.05.
>
> In short: once you know what someone bought, knowing how satisfied they were tells
> you essentially nothing further about how much they spent.

**Marking guidance.** The Rating interpretation is the exercise, and it is where most
students lose marks. Three things to look for:

1. The phrase **"holding quantity and unit price constant"** or an equivalent. A
   student who omits the conditioning has not understood multiple regression.
2. **Not** treating p = 0.058 as a clean null result. "Rating is not significant so it
   has no effect" overstates what a marginal p-value supports.
3. **Not** stating a causal claim. "Higher ratings do not lead to higher spending"
   asserts something the model cannot establish.

The strongest answers note that the sign is *negative*, which is the opposite of the
intuition most people bring, and then correctly decline to make anything of it because
the effect is too small to distinguish from noise.

::: {.callout-caution title="The point worth making about R-squared here"}
R-squared above 0.9 looks like a triumph. Ask the class what the model has actually
taught them about the business.

The answer is nothing. `Total` is `Quantity` times `Unit price` times 1.05, so the
model has rediscovered a definition. A near-perfect fit obtained by predicting a
variable from its own components is a warning sign, not an achievement.

This connects directly to the multicollinearity example in section 6 of the student
notebook, and it is the single most transferable idea in this section: **always ask
whether your predictors are logically prior to your outcome, or merely arithmetically
related to it.**
:::

# Exercise 3 - Diagnostics

```{r sol_ex3, fig.width=9, fig.height=7}
par(mfrow = c(2, 2))
plot(m2)
par(mfrow = c(1, 1))
```

```{r sol_ex3_gg}
aug2 <- augment(m2)

ggplot(aug2, aes(x = .fitted, y = .resid)) +
  geom_point(colour = "#A83296", alpha = 0.35) +
  geom_hline(yintercept = 0, colour = "#1A1A1A", linewidth = 0.8) +
  geom_smooth(se = FALSE, colour = "#D64545") +
  labs(title = "Residuals against fitted values",
       x = "Fitted value", y = "Residual") +
  theme_minimal()
```

**Expected pattern.** The residuals are tiny relative to the fitted values, and they
show a clear **funnel shape** - spread widening as the fitted value grows.

**Model answer.**

> **Linearity:** the residuals-versus-fitted plot shows no curvature, so a linear form
> is appropriate. This is expected, since the true relationship is exactly linear by
> construction.
>
> **Independence:** each row is a separate transaction, so independence is reasonable
> on design grounds. Nothing in the data suggests otherwise.
>
> **Homoscedasticity:** this assumption **fails**. The residuals fan out as fitted
> values increase, which the Scale-Location plot confirms with an upward slope. Larger
> transactions have larger errors.
>
> **Normality of residuals:** the Q-Q plot shows departures in the tails, though with
> 1,000 observations the model is fairly robust to this.
>
> The heteroscedasticity is the finding that would make me cautious. It does not bias
> the coefficients, but it makes the standard errors - and therefore every p-value and
> confidence interval in the output - unreliable.

**Marking guidance.** Students must comment on **all four** assumptions and must
identify the funnel shape. Simply printing the plots earns nothing.

The distinction worth rewarding: heteroscedasticity affects the **standard errors**, not
the **coefficient estimates**. A student who says "the model is wrong" has overstated
it; a student who says "the coefficients are still fine but I would not trust the
p-values" has understood it.

**Why the funnel appears here.** The total is a product of two quantities, so
rounding and variation scale with the size of the transaction rather than staying
constant. Multiplicative processes routinely produce this pattern, and the standard
remedy - modelling `log(Total)` instead - is worth mentioning to stronger students even
though it is beyond the syllabus.

# Exercise 4 - Prediction

```{r sol_ex4}
new_basket <- tibble(
  Quantity     = 6,
  `Unit price` = 45,
  Rating       = 8
)

predict(m2, newdata = new_basket, interval = "confidence") |> round(2)

predict(m2, newdata = new_basket, interval = "prediction") |> round(2)
```

**Expected results.** A point prediction of **286.80**, with a confidence interval of
roughly **[280.5, 293.1]** and a prediction interval of roughly **[126.8, 446.8]** -
about 25 times wider.

Note that the pure arithmetic answer would be 6 times 45 times 1.05, or 283.5. The
model returns 286.80 because it also carries an intercept and the small negative
`Rating` term. Students who notice that gap and can explain it are doing well.

**Model answer.**

> The predicted total is about 287. The confidence interval is narrow because it
> describes the *average* total across all baskets matching this description, and
> averages are estimated precisely with 1,000 observations. The prediction interval is
> wider because it must also account for the scatter of individual transactions around
> that average.
>
> A shop manager forecasting the revenue of one particular customer should use the
> **prediction interval**. The confidence interval would be appropriate only for a
> question about the average across many such baskets - for example, budgeting for a
> thousand similar transactions.

**Marking guidance.** The choice between the intervals is the mark, and the
justification must turn on **individual versus average**. This is a genuinely common
professional error: quoting a confidence interval when a client asks about their own
case understates the uncertainty substantially, and the promise will be broken far
more often than the stated confidence level implies.

# Exercise 5 - Model comparison and judgement

```{r sol_ex5}
mod_a <- lm(Total ~ Quantity, data = sales)
mod_b <- lm(Total ~ Quantity + `Unit price`, data = sales)
mod_c <- lm(Total ~ Quantity + `Unit price` + Rating + Branch + `Customer type`,
            data = sales)

bind_rows(
  glance(mod_a) |> mutate(model = "A: Quantity"),
  glance(mod_b) |> mutate(model = "B: + Unit price"),
  glance(mod_c) |> mutate(model = "C: + Rating, Branch, Customer type")
) |>
  select(model, r.squared, adj.r.squared, sigma, df)
```

**Expected pattern.**

| Model | R-squared | Adjusted R-squared |
|---|---|---|
| A: Quantity | 0.4977 | 0.4972 |
| B: + Unit price | 0.8901 | 0.8899 |
| C: + Rating, Branch, Customer type | 0.8907 | 0.8901 |

A large jump from A to B. From B to C, plain R-squared rises by 0.0006 while adjusted
R-squared moves by almost nothing - the textbook signature of predictors that do not
earn their place.

**Model answer.**

> Adding unit price to the quantity-only model transforms the fit, taking R-squared
> from 0.498 to 0.890. Adding rating, branch and customer type produces essentially no
> further improvement: R-squared rises by 0.0006 and adjusted R-squared is unchanged to
> three decimal places, confirming those three predictors earn nothing.
>
> I would recommend **Model B** on parsimony grounds - it achieves the same explanatory
> power with three fewer predictors.
>
> However, no model here tells a manager anything actionable. `Total` is defined as
> quantity times unit price plus tax, so Model B has recovered an accounting identity
> rather than discovered a business relationship. The genuinely useful finding is the
> **negative** one: branch, customer type and satisfaction contribute nothing to
> explaining transaction value once you know what was in the basket.
>
> To model something a manager could act on, I would need a different outcome
> variable - repeat visit rate, or basket size itself - and predictors that are
> genuinely prior to it rather than components of it.

**Marking guidance.** This exercise separates students who can operate the software
from students who can think.

| Band | What it looks like |
|---|---|
| Excellent | Compares adjusted R-squared correctly, recommends a model on parsimony, **and** recognises the circularity of predicting Total from its own components. Proposes a better-posed question. |
| Good | Correct comparison and a justified recommendation, but treats the high R-squared as a genuine success. |
| Adequate | Produces the table and picks the highest R-squared without discussing parsimony. |
| Weak | Reports plain R-squared only, or recommends Model C because "it has the highest R-squared". |

The last row is the trap the exercise is built around. A student who recommends Model C
on plain R-squared has ignored the section's explicit warning that R-squared never
decreases when predictors are added.

# Teaching notes

**Where students get stuck.**

1. **Interpreting coefficients without the units.** "The slope is 2.7" is not an
   interpretation. Insist on "each additional item is associated with about 2.7 more in
   gross income" every time.
2. **Forgetting "holding the others constant"** in multiple regression. This is the
   defining feature of the technique and the thing most likely to be missing from an
   exam answer.
3. **Treating a high R-squared as proof.** This dataset is the ideal cure, because the
   best-fitting model is also the emptiest.
4. **Confusing confidence and prediction intervals.** The individual-versus-average
   framing works better than any formula.
5. **Reading diagnostic plots.** Students see four plots and do not know what "good"
   looks like. Show a clean model and a badly violated one side by side - the contrast
   teaches faster than the description does.

**Timing.** Sections 1 to 4 fill one 90-minute session; sections 5 and 6 fill a second.
Do not cut section 5 - a regression whose assumptions were never checked is not an
analysis, and diagnostics are what separates this from a curve-fitting exercise.

**A better dataset for the final project.** Because `Total` is arithmetically
determined here, the superstore dataset in `R-Version/data/` is a stronger choice for
assessed regression work. `Profit` there is genuinely predictable-but-not-determined
from `Sales`, `Discount` and `Quantity`, and the relationship with `Discount` is
strongly negative and commercially interesting. It also contains real
heteroscedasticity and influential outliers, so the diagnostics have something to find.

**Connecting to the Excel track.** Excel's Analysis ToolPak regression produces
coefficients, standard errors, t-statistics, p-values and R-squared - broadly the same
table as `summary()`. It offers residual plots but no Cook's distance, no leverage
diagnostics, and no straightforward way to produce prediction intervals. For anything
beyond a single fitted line, R is substantially less work.
