---
title: "2. Descriptive Statistics - Solutions"
subtitle: "307102 Descriptive Statistics for Business | University of Petra"
code-fold: true
code-summary: "Show the code"
---


::: {.callout-caution title="Instructor copy"}
Every chunk here runs, and every numeric result quoted in the commentary has been
verified against the standard 1,000-row `supermarket_sales.xlsx`. If your copy of
the file has been edited, re-run the notebook and adjust the quoted figures.

Section headings match the student notebook so the two can be read side by side.
:::

```{r load}
library(tidyverse)
library(readxl)
library(moments)

sales <- read_excel("../data/supermarket_sales.xlsx")
```

---

# Inline Exercises

### Section 2 - Frequency table of Customer type

```{r sol_freq}
sales |>
  count(`Customer type`) |>
  mutate(percent = round(n / sum(n) * 100, 1))
```

**Expected pattern.** Members and Normal customers split close to 50/50.

**Discussion point worth raising.** A near-even split is itself a finding. If the
chain runs a membership programme, roughly half of transactions being non-member
means half of all sales generate no customer-level data at all. Everything the
business "knows" about its customers is therefore based on a self-selected half -
and members are, almost by definition, not typical shoppers. This is a good early
introduction to **selection bias**, which returns in Part 3.

### Section 3 - Histogram of Rating

```{r sol_hist}
ggplot(sales, aes(x = Rating)) +
  geom_histogram(bins = 20, fill = "#A83296", colour = "white") +
  labs(title = "Distribution of customer ratings",
       x = "Rating", y = "Number of customers") +
  theme_minimal()
```

**Expected pattern.** Not a bell curve. The distribution is broadly flat across the
middle of the scale with a fairly hard floor around 4 and ceiling around 10.

**Why this is a useful teaching moment.** Students arrive expecting everything to be
normally distributed, because every textbook picture is. Confronting a flat,
bounded distribution in week one prevents that assumption from hardening. It also
sets up section 7 - when the Q-Q plot for `Rating` bends away at both ends,
students have already seen why.

If a class discussion opens up here, the useful question is: *what would ratings
look like if customers only bothered to leave one when they were delighted or
furious?* That would give a U-shape, which is what many real review platforms
actually show. The fact that this dataset does not look like that suggests ratings
were collected systematically rather than volunteered.

### Section 4 - Mean and median Rating by product line

```{r sol_central}
sales |>
  group_by(`Product line`) |>
  summarise(
    transactions  = n(),
    mean_rating   = round(mean(Rating), 2),
    median_rating = round(median(Rating), 2)
  ) |>
  arrange(desc(mean_rating))
```

**Expected pattern.** All six product lines land within roughly half a point of one
another, somewhere near 7.

**The point of the exercise.** Students will want to declare a "winner". Push back:
a gap of 0.3 on a 10-point scale, across groups of about 160 transactions each, is
almost certainly not a real difference. Ask them what they would need in order to
claim it was - and let the answer be unsatisfying for now, because that is exactly
what Part 4 supplies.

Note also that mean and median are close for every product line here, which is what
you expect from a roughly symmetric, bounded variable. Contrast with `Total`, where
the two diverge noticeably.

### Section 6 - IQR outlier check on Rating

```{r sol_outliers}
r_q1  <- quantile(sales$Rating, 0.25)
r_q3  <- quantile(sales$Rating, 0.75)
r_iqr <- IQR(sales$Rating)

tibble(
  q1          = r_q1,
  q3          = r_q3,
  iqr         = r_iqr,
  lower_bound = r_q1 - 1.5 * r_iqr,
  upper_bound = r_q3 + 1.5 * r_iqr
)

sales |>
  filter(Rating < r_q1 - 1.5 * r_iqr | Rating > r_q3 + 1.5 * r_iqr) |>
  nrow()
```

**Expected result: zero outliers.**

**This is the most valuable exercise in the section**, precisely because the answer
is nothing. Students who predicted "a few" need to understand why they were wrong.

Q1 is 5.5 and Q3 is 8.5, so the IQR is 3 and the bounds come out at **1 and 13**.
The ratings themselves only run from **4 to 10**. The flagging thresholds sit
outside the range the scale can even produce, so no value could ever be flagged.
A bounded variable with a wide interquartile range cannot generate IQR outliers.
It is structurally impossible.

Two conclusions to draw out:

1. **"No outliers" does not mean "clean data".** It can simply mean the variable
   has no room to produce extreme values. A student who reports "I checked for
   outliers and found none" without understanding this has learned a ritual, not a
   method.
2. **The 1.5 x IQR rule is a convention, not a law.** The 1.5 was chosen by John
   Tukey because it worked reasonably for roughly normal data. It has no special
   authority, and it behaves quite differently on bounded, skewed or discrete
   variables.

Contrast with `Total`, where the same rule does flag a handful of points - and
where, as the notebook argues, almost none of them should be deleted. The lower
bound for `Total` comes out negative, so only the high tail can ever be flagged.
That asymmetry is worth pointing out: on a right-skewed variable the IQR rule is
structurally one-sided.

### Section 8 - Correlation between Unit price and Rating

```{r sol_assoc}
cor(sales$`Unit price`, sales$Rating)

ggplot(sales, aes(x = `Unit price`, y = Rating)) +
  geom_point(colour = "#A83296", alpha = 0.4) +
  labs(title = "Unit price against customer rating",
       x = "Unit price", y = "Rating") +
  theme_minimal()
```

**Expected pattern.** A correlation very close to zero, and a scatterplot that is a
formless cloud.

**Why the plot matters even when the number is boring.** The notebook's warning is
that a near-zero correlation could hide a curved relationship. Here it does not -
the plot confirms there is genuinely nothing there. That confirmation is the point.
Students should form the habit of plotting *before* concluding, not only when the
number looks interesting.

**The business reading.** How expensive the individual items were has no bearing on
how satisfied the customer was. That is worth knowing: it argues against the
intuition that premium products buy goodwill, and points attention elsewhere -
service, queue length, stock availability - for anything that might actually move
satisfaction.

---

# Exercise 1 - Branch Performance Report

```{r sol_ex1}
sales |>
  group_by(Branch) |>
  summarise(
    transactions  = n(),
    total_revenue = round(sum(Total), 2),
    mean_sale     = round(mean(Total), 2),
    median_sale   = round(median(Total), 2),
    sd_sale       = round(sd(Total), 2),
    cv_percent    = round(sd(Total) / mean(Total) * 100, 1)
  ) |>
  arrange(desc(total_revenue))
```

**Expected pattern.** Branch C leads on revenue (about 110,600) with A and B
essentially tied (about 106,200 each - within two currency units of one another,
which is worth pointing out as a coincidence rather than a finding). Transaction
counts run 328 to 340. Mean sale runs about 312 (A) to 337 (C); standard deviations
run about 232 to 263.

**Marking guidance.** The table is the easy part. The marks should be for the
interpretation, and specifically for whether the student notices that **the
differences are small relative to the spread**. Branch C's mean is about 25 above
Branch A's, but the standard deviation within each branch is around 245 - so the
gap between branches is roughly a tenth of the variation inside one branch.

A strong answer says something like: "Branch C leads on total revenue, but the gap
is about a tenth of one standard deviation and could easily be chance. On this
evidence I would not treat any branch as a better performer."

A weak answer ranks the branches and stops.

**Extension for stronger students.** Ask why `mean_sale` exceeds `median_sale` in
every branch, and what that consistency implies. The answer - that right skew is a
property of retail transaction data generally, not of any one branch - is a genuine
insight.

# Exercise 2 - Product Line Profitability

```{r sol_ex2}
sales |>
  group_by(`Product line`) |>
  summarise(
    transactions   = n(),
    total_income   = round(sum(`gross income`), 2),
    avg_income     = round(mean(`gross income`), 2),
    avg_basket     = round(mean(Total), 2)
  ) |>
  arrange(desc(total_income))
```

```{r sol_ex2b}
sales |>
  group_by(`Product line`) |>
  summarise(avg_income = round(mean(`gross income`), 2)) |>
  arrange(desc(avg_income))
```

**Expected pattern - and it works out beautifully here.** **Food and beverages**
tops total gross income (about 2,674). But **Home and lifestyle** tops average
income per transaction (about 16.0) while sitting **fifth of six** on total income,
because it has the fewest transactions after Health and beauty.

So the two questions give genuinely different answers, which is exactly the point of
the exercise.

**The concept being tested.** Total income is a product of two things: how much
each sale earns, and how many sales there are. A line can lead on total by selling
a great deal at a modest margin, or by selling little at a high one. These call for
opposite management responses:

- **High volume, low margin** - protect the volume. Do not raise prices casually,
  and make sure it is always in stock.
- **Low volume, high margin** - growth potential. Ask why volume is low. Is it
  shelf position, awareness, or genuinely limited demand?

**Marking guidance.** Full credit requires the student to compute both figures
*and* to say why the distinction changes what a manager should do. A student who
produces both columns without connecting them to a decision has done the
arithmetic but missed the analysis.

**Worth flagging in class.** In this dataset `gross income` is **exactly 4.7619
percent of `Total` for every single row** - the `gross margin percentage` column is
a constant. Margin therefore does not vary by product line at all, and the
differences in average income per transaction are purely differences in **basket
size**, not in profitability.

This is worth saying out loud, because it means the exercise's apparent finding is
partly an artefact of how the data was generated. Students can verify it in one
line:

```{r sol_ex2_margin}
range(sales$`gross income` / sales$Total * 100)
```

Recognising when a dataset cannot answer the question you are asking is a more
advanced skill than computing the answer, and this is a safe place to practise it.
A good follow-up question: *what column would we need in order to compare true
profitability across product lines?*

# Exercise 3 - Satisfaction By Payment Method

```{r sol_ex3}
sales |>
  group_by(Payment) |>
  summarise(
    transactions  = n(),
    mean_rating   = round(mean(Rating), 2),
    median_rating = round(median(Rating), 2),
    sd_rating     = round(sd(Rating), 2)
  ) |>
  arrange(desc(mean_rating))
```

```{r sol_ex3b}
ggplot(sales, aes(x = Payment, y = Rating)) +
  geom_boxplot(fill = "#A83296", alpha = 0.7) +
  labs(title = "Customer rating by payment method",
       x = "Payment method", y = "Rating") +
  theme_minimal()
```

**Expected pattern.** Credit card 7.00, Cash 6.97, Ewallet 6.95 - a total spread of
**0.05** - with standard deviations near 1.7 and three boxplots that overlap almost
completely.

**The answer to "do the differences matter?" is no**, and this is about as clear-cut
as it gets: the gap between the best and worst payment method is one thirty-fifth of
the variation within any one of them. The boxplot makes that obvious at a glance.
This is the exercise where the visual does the arguing.

**Marking guidance.** Look for the student comparing the size of the difference
between groups against the size of the variation within groups. A 0.05 gap between
groups whose internal standard deviation is 1.7 is nothing at all. Students who articulate
that comparison - even informally, as "the boxes overlap far too much" - have
grasped the intuition behind every hypothesis test in Part 4.

**Useful follow-up question for discussion.** If the business had run this analysis
on 50 transactions instead of 1,000, would they have been more or less likely to
find a "difference"? Smaller samples produce noisier group means, so more likely -
and that is precisely how spurious findings get into business reports.

# Exercise 4 - Shape Investigation: Unit Price

```{r sol_ex4}
sales |>
  summarise(
    mean_price   = round(mean(`Unit price`), 2),
    median_price = round(median(`Unit price`), 2),
    sd_price     = round(sd(`Unit price`), 2),
    skewness     = round(skewness(`Unit price`), 3),
    kurtosis     = round(kurtosis(`Unit price`), 3),
    excess_kurt  = round(kurtosis(`Unit price`) - 3, 3)
  )
```

```{r sol_ex4b}
ggplot(sales, aes(x = `Unit price`)) +
  geom_histogram(bins = 30, fill = "#A83296", colour = "white") +
  geom_vline(aes(xintercept = mean(`Unit price`)),
             colour = "#1A1A1A", linewidth = 1) +
  geom_vline(aes(xintercept = median(`Unit price`)),
             colour = "#1A1A1A", linewidth = 1, linetype = "dashed") +
  labs(title = "Unit price, with mean (solid) and median (dashed)",
       x = "Unit price", y = "Frequency") +
  theme_minimal()
```

```{r sol_ex4c}
ggplot(sales, aes(sample = `Unit price`)) +
  stat_qq(colour = "#A83296", alpha = 0.5) +
  stat_qq_line(colour = "#1A1A1A") +
  labs(title = "Q-Q plot: Unit price",
       x = "Theoretical normal quantiles", y = "Observed unit price") +
  theme_minimal()
```

**Expected pattern.** Skewness close to zero, mean and median nearly identical, a
roughly flat histogram, and a Q-Q plot with a pronounced S-shape.

**Model answer.**

> Unit price is essentially symmetric - skewness is near zero and the mean and
> median are almost the same, so neither is distorted by a tail. But it is not
> normal. The histogram is close to flat rather than bell-shaped, and the excess
> kurtosis is clearly negative, meaning far less of the data sits in the tails than
> a normal distribution would put there. The Q-Q plot confirms this with an S-shape:
> the observed values run out of room at both ends because prices are bounded at
> roughly 10 and 100. Since the distribution is symmetric, the mean and median are
> equally good summaries, and the mean is conventional.

**The trap this exercise sets.** Students frequently equate "symmetric" with
"normal". Unit price is the counterexample: symmetric, and emphatically not normal.
A uniform distribution is perfectly symmetric and has nothing in common with a bell
curve. If a student writes "skewness is near zero therefore the data is normal",
that is the misconception to correct.

**On the kurtosis figure.** Encourage students to report the excess value
(subtracting 3) so it is comparable with Excel and with published figures, but to
say which convention they are using. Unstated conventions are how numbers get
misread.

# Exercise 5 - A Question Of Your Own

Open-ended. Below are three worked examples at increasing levels of sophistication,
usable as exemplars or as a marking reference.

### Level 1 - Descriptive, competently done

*Does the busiest branch also have the happiest customers?*

```{r sol_ex5a}
sales |>
  group_by(Branch) |>
  summarise(
    transactions = n(),
    avg_rating   = round(mean(Rating), 2)
  ) |>
  arrange(desc(transactions))
```

Correct, clearly presented, but asks little of the data.

### Level 2 - Two variables combined

*Do members spend more than non-members, and is the difference consistent across
branches?*

```{r sol_ex5b}
sales |>
  group_by(Branch, `Customer type`) |>
  summarise(
    transactions = n(),
    avg_spend    = round(mean(Total), 2),
    .groups = "drop"
  ) |>
  pivot_wider(names_from = `Customer type`, values_from = c(transactions, avg_spend))
```

```{r sol_ex5b2}
ggplot(sales, aes(x = Branch, y = Total, fill = `Customer type`)) +
  geom_boxplot(alpha = 0.8) +
  scale_fill_manual(values = c("#A83296", "#D070D0")) +
  labs(title = "Spend by customer type within each branch",
       x = "Branch", y = "Transaction total") +
  theme_minimal()
```

Better. It asks whether a pattern holds up when you split it further - the
beginning of controlling for a third variable, which is what Part 5 formalises.

### Level 3 - Aware of its own limitations

*Which hour of the day generates the most revenue, and should staffing follow it?*

```{r sol_ex5c}
sales |>
  # readxl may read Time as a date-time or as text depending on the file.
  # Pulling the digits before the first colon works either way.
  mutate(hour = as.numeric(str_extract(as.character(Time), "\\d{1,2}(?=:)"))) |>
  group_by(hour) |>
  summarise(
    transactions  = n(),
    total_revenue = round(sum(Total), 2),
    avg_sale      = round(mean(Total), 2)
  ) |>
  arrange(desc(total_revenue))
```

```{r sol_ex5c2}
sales |>
  # readxl may read Time as a date-time or as text depending on the file.
  # Pulling the digits before the first colon works either way.
  mutate(hour = as.numeric(str_extract(as.character(Time), "\\d{1,2}(?=:)"))) |>
  group_by(hour) |>
  summarise(total_revenue = sum(Total)) |>
  ggplot(aes(x = hour, y = total_revenue)) +
  geom_col(fill = "#A83296") +
  labs(title = "Revenue by hour of day",
       x = "Hour", y = "Total revenue") +
  theme_minimal()
```

The strongest answers will note that revenue per hour reflects **when the shop is
open and staffed**, not raw demand - a quiet hour might be quiet because there are
no tills open. The data cannot distinguish these, so it cannot by itself justify a
staffing change. Recognising that limitation is worth more than the chart.

**Marking guidance for Exercise 5.**

| Band | What it looks like |
|---|---|
| Excellent | Non-obvious question; correct method; interprets in business terms; states what the data cannot tell you |
| Good | Sensible question; correct method; clear interpretation |
| Adequate | Question answered correctly but restates a worked example; interpretation thin |
| Weak | Code runs but no interpretation, or a conclusion the output does not support |

The most common failure is a technically flawless pipeline with no sentence saying
what it means. Weight the interpretation heavily, and say so in advance.

# Teaching Notes

**Where students get stuck, in order of frequency.**

1. **Backticks.** Column names with spaces cause more failed chunks than everything
   else combined. Say it in the first five minutes and put it on the board.
2. **`=` versus `==`.** Inside `filter()`, a single equals sign produces an error
   whose message does not obviously point at the cause.
3. **`+` versus `|>`.** Inside `ggplot()` layers join with `+`. Outside it, steps
   join with `|>`. Mixing them is near-universal in week one.
4. **Forgetting `library(tidyverse)`** after restarting R, then meeting
   `could not find function "filter"` and assuming the installation is broken.
5. **Case sensitivity.** `Rating` works; `rating` does not. Recommend running
   `names(sales)` and copying from the output.

**Timing.** Sections 1 to 4 comfortably fill one 90-minute session. Sections 5 to 8
fill a second. Section 9 works well as a short demonstration at the start of the
Part 3 session, where it lands better than at the end of Part 1.

**If you are short of time**, section 9 is the one to cut, since Part 3 covers the
same ground formally. Do not cut the Q-Q plots in section 7 - Part 4 depends on
students being able to read one.

**Connecting to the Excel track.** Students who did the Excel version may want a
mapping. The main correspondences: `mean` for `AVERAGE`, `sd` for `STDEV.S`,
`quantile` for `PERCENTILE.INC`, `cor` for `CORREL`, `count` for a PivotTable, and
`group_by` plus `summarise` for a PivotTable with a row field. Flag the skewness
and kurtosis conventions, which differ as described in the student notebook.
