---
title: "2. Descriptive Statistics"
subtitle: "307102 Descriptive Statistics for Business | University of Petra"
---

## New R this week

Five new functions. That is all you need to add to section 1.

| Function | What it does | Why a business needs it |
|---|---|---|
| `count()` | Counts rows in each category | How many sales came from each branch |
| `ggplot()` | Draws a graph | Turning numbers into something a manager will look at |
| `sd()`, `var()` | Standard deviation and variance | How unpredictable are our daily takings |
| `quantile()` | Percentiles and quartiles | What a typical and a top-10-percent customer spend |
| `cor()` | Correlation between two columns | Does a bigger basket mean a happier customer |

## Objectives

By the end of this notebook you will be able to:

- Identify the measurement level of a variable and say which statistics are legitimate for it
- Build frequency tables and read what they tell you
- Draw and interpret bar charts, histograms, boxplots and scatterplots
- Calculate and choose between mean, median and mode
- Measure spread with range, variance, standard deviation and the coefficient of variation
- Use quartiles and the IQR rule to identify outliers
- Describe distribution shape using skewness, kurtosis and Q-Q plots
- Measure association between two variables, and explain why that is not causation

## Setup

Run this chunk first, every time you open this notebook.

```{r load}
library(tidyverse)
library(readxl)
library(moments)

sales <- read_excel("../data/supermarket_sales.xlsx")

glimpse(sales)
```

A reminder of the business situation. This is a supermarket chain with three
branches. Each of the 1,000 rows is one transaction: what was bought, by whom, how
much was paid, and what rating the customer left afterwards.

Throughout, keep asking the question a manager would ask: **so what?** A number that
does not change a decision is not worth calculating.

# 1. Levels of measurement

Before calculating anything, you must know what kind of variable you are holding.
This is not box-ticking - it determines which calculations are meaningful and which
are nonsense.

| Level | What it does | Can you...? | Example here |
|---|---|---|---|
| **Nominal** | Names categories | Count them. Nothing else. | `Branch`, `Gender`, `Payment` |
| **Ordinal** | Categories with a rank order | Also order them. Gaps are not equal. | A scale from Poor to Excellent |
| **Interval** | Equal gaps, no true zero | Also subtract. Ratios are meaningless. | `Rating`, `Date` |
| **Ratio** | Has a true zero | Also divide. Ratios are meaningful. | `Total`, `Quantity`, `Unit price` |

The test separating interval from ratio is whether zero means "none of it". A
`Total` of 0 means no money changed hands, so it is ratio, and one sale can
sensibly be twice as large as another. A `Rating` of 0 does not mean "no
satisfaction" - it is just the bottom of a scale someone invented - so a rating of 8
is not "twice as satisfied" as a 4.

R enforces some of this for you:

```{r levels_demo, warning=TRUE}
mean(sales$Branch)
```

R returns `NA` and warns that the argument is not numeric. That warning is R
protecting you: the average of "A", "B" and "C" does not exist.

::: {.callout-warning title="The computer checks types; you check sense"}
R will **not** protect you from averaging something numeric that should not be
averaged. If a variable were coded 1 = Cash, 2 = Credit, 3 = Ewallet, R would
happily report a mean of 2.03.

That number is meaningless, and only you can know it.
:::

:::: {.callout-note title="Self-check 1"}
A colleague reports that "average customer type is 1.6" after coding Member as 1 and
Normal as 2. What has gone wrong, and what should they have reported instead?

::: {.callout-note collapse="true" title="Show the answer"}
`Customer type` is **nominal**. The numbers 1 and 2 are labels, not quantities - the
choice of which group gets 1 was arbitrary, and reversing it would change the
"average" to 1.4 without changing anything about the customers.

The correct summary is a **count or a percentage**: "60 percent of transactions were
from Normal customers". That statement survives relabelling; the average does not.

This mistake is common with survey data, where everything arrives as numbers
regardless of what it actually measures.
:::
::::

# 2. Frequency tables

For a nominal variable, counting is the whole analysis.

```{r freq_basic}
sales |>
  count(Branch)
```

Raw counts are hard to compare when groups differ in size, so convert to
percentages:

```{r freq_pct}
sales |>
  count(Branch) |>
  mutate(percent = round(n / sum(n) * 100, 1))
```

Sorting by size makes the story jump out, which matters with more than three
categories:

```{r freq_sorted}
sales |>
  count(`Product line`, sort = TRUE) |>
  mutate(percent = round(n / sum(n) * 100, 1))
```

## Cross-tabulation

Two nominal variables at once. This is where frequency tables earn their keep,
because it lets you ask whether one thing depends on another:

```{r freq_cross}
sales |>
  count(Branch, Payment) |>
  pivot_wider(names_from = Payment, values_from = n)
```

`pivot_wider()` reshapes the long list into a grid, which is far easier to read.

Do the three branches have noticeably different payment habits? Eyeballing a
difference is a fine first step. In section 3.3 you will learn the chi-square test,
which tells you whether a difference this size could plausibly be chance.

## Grouping a numeric variable into classes

Frequency tables also work on numeric data once you cut it into bands.

```{r freq_numeric}
sales |>
  mutate(spend_band = cut(Total,
                          breaks = c(0, 100, 250, 500, 750, 1100),
                          labels = c("Under 100", "100-250", "250-500",
                                     "500-750", "Over 750"))) |>
  count(spend_band) |>
  mutate(percent = round(n / sum(n) * 100, 1))
```

Notice how much information this throws away - 1,000 distinct values become five
bands. It also introduces a choice: those boundaries were a judgement, and different
boundaries would tell a slightly different story. Frequency tables for numeric data
are a summary for presentation, not a basis for analysis.

::: {.callout-note title="Your turn"}
Build a frequency table of `Customer type` with percentages.

```{r ex_freq, eval=FALSE}
sales |>
  count(`______`) |>
  mutate(percent = round(n / sum(n) * 100, 1))
```

:::

# 3. Visualising the data

A table of numbers is precise. A graph is persuasive. You need both.

Every graph in R follows the same three-part recipe:

1. `ggplot(data, aes(...))` - which data, and which columns map to which visual property
2. `+ geom_something()` - what to draw
3. `+ labs(...)` - titles and axis labels

::: {.callout-warning title="Plus, not pipe"}
Inside `ggplot()`, layers join with `+`. Outside it, steps join with `|>`. The two
are not interchangeable, and mixing them is near-universal in week one.
:::

## Bar chart - for categories

```{r plot_bar}
ggplot(sales, aes(x = `Product line`)) +
  geom_bar(fill = "#A83296") +
  labs(title = "Transactions by product line",
       x = NULL, y = "Number of transactions") +
  coord_flip() +
  theme_minimal()
```

`coord_flip()` turns the bars sideways. With long category names this is almost
always right - rotated vertical labels are hard to read and look amateurish.

## Histogram - for numeric data

```{r plot_hist}
ggplot(sales, aes(x = Total)) +
  geom_histogram(bins = 30, fill = "#A83296", colour = "white") +
  labs(title = "Distribution of transaction values",
       x = "Transaction total", y = "Number of transactions") +
  theme_minimal()
```

**The bin count is a real choice, not a detail.** Change it and the story changes:

```{r plot_hist_bins}
ggplot(sales, aes(x = Total)) +
  geom_histogram(bins = 5, fill = "#A83296", colour = "white") +
  labs(title = "The same data with only 5 bins",
       x = "Transaction total", y = "Number of transactions") +
  theme_minimal()
```

With 5 bins the distribution looks smooth and unremarkable. With 60 it would look
spiky. Neither is wrong, but neither is neutral. Always look at your data with two
or three bin settings before deciding what it "looks like" - and be suspicious when
someone shows you a histogram without saying how they binned it.

## Boxplot - for comparing groups

```{r plot_box}
ggplot(sales, aes(x = Branch, y = Total)) +
  geom_boxplot(fill = "#A83296", alpha = 0.7) +
  labs(title = "Transaction values by branch",
       x = "Branch", y = "Transaction total") +
  theme_minimal()
```

A boxplot compresses a whole distribution into five numbers:

- The **thick line** is the median - half the sales are above it
- The **box** spans the middle 50 percent, from the 25th to the 75th percentile
- The **whiskers** reach the most extreme values within 1.5 box-lengths of the box
- Any **dots** beyond the whiskers are flagged as potential outliers

Boxplots are the fastest way to compare several groups at once. Look at whether the
boxes overlap: heavily overlapping boxes mean the branches are more similar than
different, whatever the exact averages say.

## Scatterplot - for two numeric variables

```{r plot_scatter}
ggplot(sales, aes(x = Quantity, y = Total)) +
  geom_point(colour = "#A83296", alpha = 0.4) +
  labs(title = "Basket size against transaction value",
       x = "Items purchased", y = "Transaction total") +
  theme_minimal()
```

`alpha = 0.4` makes each point semi-transparent, so where many points overlap the
colour builds up. With 1,000 points this reveals density that solid dots hide.

## Adding a third variable with colour

```{r plot_scatter_colour}
ggplot(sales, aes(x = Quantity, y = Total, colour = Branch)) +
  geom_point(alpha = 0.5) +
  labs(title = "Basket size against value, by branch",
       x = "Items purchased", y = "Transaction total") +
  scale_colour_manual(values = c("#A83296", "#D070D0", "#5C1A54")) +
  theme_minimal()
```

::: {.callout-note title="Your turn"}
Draw a histogram of `Rating` with 20 bins.

```{r ex_plot, eval=FALSE}
ggplot(sales, aes(x = ______)) +
  geom_histogram(bins = ____, fill = "#A83296", colour = "white") +
  labs(title = "Distribution of customer ratings",
       x = "Rating", y = "Number of customers") +
  theme_minimal()
```

Look at the result. Does it have the bell shape you might have expected, or
something else? Hold that thought - section 7 comes back to it.
:::

# 4. Measures of central tendency

Three ways to answer "what is typical?", and they do not always agree.

## Mean

```{r central_mean}
mean(sales$Total)
```

The arithmetic average. It uses every value, which is its strength, and it is
dragged by extreme values, which is its weakness.

## Median

```{r central_median}
median(sales$Total)
```

The middle value when everything is lined up in order. Half the transactions are
above it and half below. It ignores how far away the extremes are, which makes it
**robust** - resistant to outliers.

## Mode

::: {.callout-warning title="There is no mode() function for this"}
R does have a function called `mode()`, but it reports what *type* an object is -
a completely different thing, and a genuine source of confusion.
:::

For a categorical variable, the mode is the top row of a sorted count:

```{r central_mode_cat}
sales |>
  count(`Product line`, sort = TRUE) |>
  slice(1)
```

For a continuous variable like `Total`, the mode is not useful - with 1,000
transactions recorded to two decimal places, almost every value appears once. The
mode earns its place with categorical data and with discrete counts:

```{r central_mode_num}
sales |>
  count(Quantity, sort = TRUE) |>
  slice(1:3)
```

## Why the choice matters

```{r central_compare}
sales |>
  summarise(
    mean_total   = mean(Total),
    median_total = median(Total),
    difference   = mean(Total) - median(Total)
  )
```

The mean sits above the median. That gap is not noise - it tells you the
distribution has a longer tail on the high side. A few very large transactions pull
the average up while the median stays put.

Here is why that matters commercially. If you report the mean as your "average
transaction value" target, a handful of unusually large baskets can make performance
look better than the typical customer experience actually is. The median answers a
different and often more useful question: what does a normal shopping trip look
like?

```{r central_robust}
totals_with_error <- c(sales$Total, 500000)   # one data entry error

tibble(
  statistic = c("Mean", "Median"),
  original  = c(mean(sales$Total), median(sales$Total)),
  with_typo = c(mean(totals_with_error), median(totals_with_error))
)
```

One mistyped figure moves the mean noticeably. The median barely notices. When you
suspect your data has errors - and real business data always does - the median is
the safer summary.

:::: {.callout-note title="Self-check 2"}
A property firm reports that the mean house price in a neighbourhood is 480,000 and
the median is 310,000. What does that gap tell you, and which figure should a
first-time buyer pay attention to?

::: {.callout-note collapse="true" title="Show the answer"}
The mean sitting far above the median means the distribution is **right-skewed**: a
small number of very expensive properties are pulling the average up, while most
houses cluster much lower.

A first-time buyer should look at the **median**, and ideally the 25th percentile
too. The median describes the house in the middle of the market. The mean describes
a house that may not exist anywhere in the neighbourhood.

This is why property and income statistics are almost always reported as medians.
Any time you see a mean quoted for something with a natural upper tail - salaries,
house prices, insurance claims, transaction values - ask what the median is.
:::
::::

## Comparing groups

```{r central_groups}
sales |>
  group_by(Branch) |>
  summarise(
    transactions = n(),
    mean_total   = round(mean(Total), 2),
    median_total = round(median(Total), 2)
  )
```

::: {.callout-note title="Your turn"}
Find the mean and median `Rating` for each product line, sorted from highest mean
down.

```{r ex_central, eval=FALSE}
sales |>
  group_by(`______`) |>
  summarise(
    mean_rating   = round(mean(______), 2),
    median_rating = round(median(______), 2)
  ) |>
  arrange(desc(______))
```

:::

# 5. Measures of dispersion

Two branches can have identical average sales and behave completely differently. The
average tells you where the centre is; dispersion tells you how far things scatter
around it.

This matters more than students expect. A supplier who delivers in 5 days on
average, every time, is a completely different business partner from one who
averages 5 days but ranges from 1 to 20.

## Range

```{r disp_range}
range(sales$Total)

max(sales$Total) - min(sales$Total)
```

Simple, and almost useless. It depends entirely on two values - the two most likely
to be errors.

## Variance and standard deviation

```{r disp_var_sd}
var(sales$Total)

sd(sales$Total)
```

The variance is the average squared distance from the mean. Squaring makes it work
mathematically, but it also means the variance is in *squared currency units*, which
nobody can interpret.

The standard deviation is its square root, which puts it back into the original
units. That is why you almost always report the standard deviation and almost never
the variance - though the variance reappears in section 3.3, where the squaring
becomes genuinely useful.

Read the standard deviation as **the typical distance a transaction sits from the
average transaction**.

## Coefficient of variation

Which varies more - transaction totals, or customer ratings?

You cannot compare their standard deviations directly. One is currency running into
the hundreds; the other is a 1-to-10 scale. The **coefficient of variation**
expresses the standard deviation as a percentage of the mean, cancelling the units:

```{r disp_cv}
sales |>
  summarise(
    cv_total    = round(sd(Total)   / mean(Total)    * 100, 1),
    cv_rating   = round(sd(Rating)  / mean(Rating)   * 100, 1),
    cv_quantity = round(sd(Quantity)/ mean(Quantity) * 100, 1)
  )
```

Now the comparison is fair. A higher CV means more relative variability.

The result is intuitive once you see it: ratings are constrained to a narrow 1-to-10
band and cluster near the middle, so they vary little in relative terms. Transaction
totals can be almost anything, so they vary a great deal.

::: {.callout-warning title="When the CV is meaningless"}
The CV only works for ratio variables with a true zero and a positive mean. Applying
it to temperatures in Celsius, or to a scale that could sit anywhere on the number
line, produces a number with no interpretation.
:::

## Comparing consistency between groups

```{r disp_groups}
sales |>
  group_by(Branch) |>
  summarise(
    mean_total = round(mean(Total), 2),
    sd_total   = round(sd(Total), 2),
    cv_percent = round(sd(Total) / mean(Total) * 100, 1)
  )
```

Two branches with similar means but different standard deviations are running
different businesses. The one with lower spread has more predictable revenue, which
makes staffing and stock planning easier - a real operational advantage the average
completely hides.

:::: {.callout-note title="Self-check 3"}
Two investment funds both returned 8 percent on average over ten years. Fund A had a
standard deviation of 3 percent; Fund B, 19 percent. What does that tell you, and is
one straightforwardly better?

::: {.callout-note collapse="true" title="Show the answer"}
Fund A's returns clustered tightly around 8 percent every year. Fund B's swung wildly
- it likely had years of large gains and years of heavy losses that happened to
average out.

**Neither is straightforwardly better**, and that is the point. Fund B carries far
more **risk**, which matters enormously if you might need the money at a bad moment,
and matters less over a thirty-year horizon. An investor near retirement should
strongly prefer A; a young investor might rationally accept B's volatility.

In finance, the standard deviation of returns is the standard definition of risk. So
this is not a case of the average being wrong - it is a case of the average
answering only half the question.
:::
::::

# 6. Percentiles, quartiles and outliers

A **percentile** is the value below which a given percentage of the data falls. The
75th percentile is the value 75 percent of transactions come in under.

```{r pct_quantile}
quantile(sales$Total)
```

Those five numbers are the **five-number summary**, and they are exactly what a
boxplot draws. The 25th, 50th and 75th percentiles are the first, second and third
**quartiles** because they cut the data into four equal parts.

You can ask for any percentile:

```{r pct_custom}
quantile(sales$Total, probs = c(0.10, 0.50, 0.90, 0.99))
```

The 99th percentile is worth pausing on. It tells you what your top 1 percent of
transactions look like - useful when deciding whether a "big spender" programme is
worth running, and where to set its threshold.

## The interquartile range

```{r pct_iqr}
IQR(sales$Total)
```

The IQR is the width of the middle 50 percent: Q3 minus Q1. Like the median it is
robust - the extreme values that inflate the range and the standard deviation do not
touch it.

## The 1.5 x IQR rule

```{r pct_outlier_bounds}
q1  <- quantile(sales$Total, 0.25)
q3  <- quantile(sales$Total, 0.75)
iqr <- IQR(sales$Total)

lower_bound <- q1 - 1.5 * iqr
upper_bound <- q3 + 1.5 * iqr

tibble(q1, q3, iqr, lower_bound, upper_bound)
```

```{r pct_outliers}
outliers <- sales |>
  filter(Total < lower_bound | Total > upper_bound)

nrow(outliers)
```

The `|` means **or**: keep a row if it is below the lower bound *or* above the upper
bound.

```{r pct_outliers_show}
outliers |>
  select(`Invoice ID`, Branch, `Product line`, Quantity, Total) |>
  arrange(desc(Total))
```

## What to do with an outlier

::: {.callout-tip title="Flagging is not deleting"}
**The 1.5 x IQR rule identifies unusual values. It does not identify wrong ones.**
Deleting points because a rule flagged them is one of the most common ways to
quietly corrupt an analysis.
:::

Work through three questions:

1. **Is it possible?** A transaction total of -50, or a customer age of 210, is an
   error. Investigate and correct or remove it.
2. **Is it a different thing?** A supermarket transaction of 30,000 might be a
   legitimate wholesale order that does not belong in an analysis of retail
   shoppers. Analyse it separately rather than deleting it.
3. **Is it just large?** Rich customers exist. A genuine, correctly recorded large
   sale is real data, and removing it because it is inconvenient is falsifying your
   results.

Only category 1 justifies deletion. Whatever you decide, **write down what you did
and why** - an analysis where points vanished without explanation cannot be trusted
by anyone, including you in six months.

::: {.callout-note title="Your turn"}
Apply the same IQR outlier check to `Rating`. Before running it, predict how many
outliers you will find.

```{r ex_outliers, eval=FALSE}
r_q1  <- quantile(sales$Rating, ____)
r_q3  <- quantile(sales$Rating, ____)
r_iqr <- IQR(sales$______)

sales |>
  filter(Rating < r_q1 - 1.5 * r_iqr | Rating > r_q3 + 1.5 * r_iqr) |>
  nrow()
```

Was your prediction right? If the answer surprised you, think about what the
histogram of `Rating` looked like in section 3.
:::

# 7. The shape of a distribution

Centre and spread do not fully describe a distribution. Two datasets can share both
and look nothing alike. Shape is the third piece.

## Skewness

Skewness measures lopsidedness.

- **Positive (right) skew**: a long tail to the high side. Mean above median.
- **Negative (left) skew**: a long tail to the low side. Mean below median.
- **Zero**: symmetric.

```{r shape_skew}
sales |>
  summarise(
    skew_total    = round(skewness(Total), 3),
    skew_rating   = round(skewness(Rating), 3),
    skew_quantity = round(skewness(Quantity), 3)
  )
```

Rough guide: below 0.5 in absolute value is fairly symmetric; 0.5 to 1 is moderately
skewed; above 1 is strongly skewed.

Now check the number agrees with the picture, which you should always do:

```{r shape_skew_plot}
ggplot(sales, aes(x = Total)) +
  geom_histogram(bins = 30, fill = "#A83296", colour = "white") +
  geom_vline(aes(xintercept = mean(Total)),
             colour = "#1A1A1A", linewidth = 1) +
  geom_vline(aes(xintercept = median(Total)),
             colour = "#1A1A1A", linewidth = 1, linetype = "dashed") +
  labs(title = "Transaction totals, with mean (solid) and median (dashed)",
       x = "Transaction total", y = "Number of transactions") +
  theme_minimal()
```

The solid mean line sitting to the right of the dashed median line is the visual
signature of right skew.

## Kurtosis

Kurtosis measures how much of the distribution lives in the tails - how prone it is
to extreme values.

```{r shape_kurt}
sales |>
  summarise(
    kurt_total  = round(kurtosis(Total), 3),
    kurt_rating = round(kurtosis(Rating), 3)
  )
```

::: {.callout-warning title="Important if you have used Excel"}
`moments::kurtosis()` returns **raw kurtosis**, where a perfect normal distribution
scores **3**. Excel's `KURT()` returns **excess kurtosis**, where a normal
distribution scores **0** - it has already subtracted the 3.

To compare with an Excel result, subtract 3: `kurtosis(sales$Total) - 3`

Excel and `moments` also use slightly different sample corrections for skewness, so
those figures differ in the second or third decimal place. The interpretation is
unaffected, but do not be alarmed by a small discrepancy when checking your Excel
work against your R work.
:::

## Q-Q plots - a better test of normality

Skewness and kurtosis compress the whole shape into two numbers. A **Q-Q plot** shows
the entire comparison at once, plotting your data against what a perfect normal
distribution would produce. If the data were normal, the points would sit on the
diagonal line.

```{r shape_qq_rating}
ggplot(sales, aes(sample = Rating)) +
  stat_qq(colour = "#A83296", alpha = 0.5) +
  stat_qq_line(colour = "#1A1A1A") +
  labs(title = "Q-Q plot: Rating",
       x = "Theoretical normal quantiles", y = "Observed ratings") +
  theme_minimal()
```

```{r shape_qq_total}
ggplot(sales, aes(sample = Total)) +
  stat_qq(colour = "#A83296", alpha = 0.5) +
  stat_qq_line(colour = "#1A1A1A") +
  labs(title = "Q-Q plot: Total",
       x = "Theoretical normal quantiles", y = "Observed totals") +
  theme_minimal()
```

Reading a Q-Q plot:

- Points **on the line** - close to normal
- Points curving **up at the right end** - a heavy right tail, so right skew
- Points curving **away at both ends** - heavier tails than normal
- A distinct **S-shape** - lighter tails than normal

::: {.callout-tip title="Why this matters beyond section 2"}
Several tests in section 3.3 assume the data is roughly normal, and a Q-Q plot is
the quickest honest way to check that before you rely on it.

Learning to read one now will save you from a confidently-reported wrong answer
later.
:::

:::: {.callout-note title="Self-check 4"}
Why might a distribution of customer ratings on a 1-to-10 scale fail a normality
check even when nothing is wrong with the data?

::: {.callout-note collapse="true" title="Show the answer"}
Because it is **bounded and discrete**. A normal distribution runs from minus
infinity to plus infinity and takes any value in between. A rating scale stops hard
at its endpoints and only lands on a limited set of values.

That produces two visible effects on a Q-Q plot: a staircase pattern from the
discreteness, and the ends bending away from the line because the data cannot run
off to infinity the way the normal distribution expects.

The wider lesson: **"not normal" is not the same as "bad data"**. Ratings, counts,
waiting times and proportions are all routinely non-normal for perfectly good
structural reasons. The question is never "is this normal?" but "is this close
enough to normal for the method I want to use?" - and section 3.3 gives you the
tools to answer that properly.
:::
::::

# 8. Measures of association

So far, one variable at a time. Now: do two variables move together?

## Covariance

```{r assoc_cov}
cov(sales$Quantity, sales$Total)
```

Positive means they tend to rise together. But the size is uninterpretable - it
depends on the units of both variables. Measure the total in fils instead of dinars
and the covariance changes by a factor of a thousand while the relationship is
identical.

## Correlation

Correlation fixes this by standardising covariance onto a fixed scale from -1 to +1:

```{r assoc_cor}
cor(sales$Quantity, sales$Total)
```

| Value | Meaning |
|---|---|
| +1 | Perfect positive relationship |
| +0.7 to +0.9 | Strong positive |
| +0.4 to +0.6 | Moderate positive |
| 0 | No **linear** relationship |
| -0.4 to -0.6 | Moderate negative |
| -1 | Perfect negative |

## A correlation matrix

```{r assoc_matrix}
sales |>
  select(`Unit price`, Quantity, Total, `gross income`, Rating) |>
  cor() |>
  round(3)
```

Read across a row and down a column to find the correlation between any pair. The
diagonal is 1 because every variable correlates perfectly with itself.

Two things here are worth noticing.

**Some correlations are near-perfect by construction, not by discovery.** `Total` and
`gross income` are both calculated from the same underlying sale, so of course they
move together. A correlation of 1.0 between two variables usually means one is a
rescaled version of the other - arithmetic rather than insight.

**Others are close to zero.** Whatever drives customer satisfaction, basket size is
not obviously it - itself a useful finding, and one a manager assuming "big spenders
are happy customers" would want to know.

## Correlation is not causation

A strong correlation permits exactly three explanations, and you cannot tell which
from the correlation alone:

1. **A causes B.** Buying more items causes a higher total. True here, and known from
   the arithmetic, not from the correlation.
2. **B causes A.** The direction can easily run the other way.
3. **Something else causes both.** Ice cream sales correlate with drowning deaths.
   Neither causes the other; hot weather causes both. This is a **confounding
   variable**, and it is the usual explanation for surprising correlations.

There is a fourth possibility that is easy to forget: **coincidence**. Search enough
pairs of variables and you will find strong correlations that mean nothing.

## Correlation only sees straight lines

`cor()` measures **linear** association. A correlation near zero does not mean "no
relationship" - it means "no straight-line relationship".

```{r assoc_nonlinear}
demo <- tibble(
  x = seq(-10, 10, length.out = 200),
  y = x^2
)

cor(demo$x, demo$y)
```

```{r assoc_nonlinear_plot}
ggplot(demo, aes(x = x, y = y)) +
  geom_point(colour = "#A83296") +
  labs(title = "Correlation is about zero, but y is completely determined by x",
       x = "x", y = "y") +
  theme_minimal()
```

The correlation is essentially zero. Yet `y` is perfectly predictable from `x` - you
could not ask for a stronger relationship.

::: {.callout-tip title="The rule"}
**Always plot the data before trusting a correlation coefficient.** A single number
cannot tell you the shape of a relationship, and it will not warn you when it has
missed one.
:::

::: {.callout-note title="Your turn"}
Find the correlation between `Unit price` and `Rating`, then plot it to check the
number is telling the truth.

```{r ex_assoc, eval=FALSE}
cor(sales$`______`, sales$______)

ggplot(sales, aes(x = `______`, y = ______)) +
  geom_point(colour = "#A83296", alpha = 0.4) +
  labs(title = "Unit price against customer rating") +
  theme_minimal()
```

:::

# 9. A look ahead - how sure are we?

Everything so far has described the 1,000 transactions in front of you. But nobody
really cares about those 1,000 transactions. They care about the business those
transactions came from.

The average transaction value you calculated is a fact about your sample. Collect a
different 1,000 transactions next month and you will get a different number. So how
much should you trust the one you have?

Here is a way to find out using nothing but the data itself. Take your 1,000
transactions, draw 1,000 of them **at random with replacement**, and compute the
mean. Do it two thousand times.

```{r bootstrap}
set.seed(2024)

boot_means <- replicate(2000, {
  resample <- sample(sales$Total, size = nrow(sales), replace = TRUE)
  mean(resample)
})

quantile(boot_means, probs = c(0.025, 0.975))
```

That interval is where the true average transaction value plausibly sits.

```{r bootstrap_plot}
tibble(boot_mean = boot_means) |>
  ggplot(aes(x = boot_mean)) +
  geom_histogram(bins = 40, fill = "#A83296", colour = "white") +
  geom_vline(xintercept = mean(sales$Total),
             colour = "#1A1A1A", linewidth = 1) +
  labs(title = "2,000 resampled averages; the solid line is our actual sample mean",
       x = "Resampled mean transaction value", y = "Frequency") +
  theme_minimal()
```

Two things about that histogram deserve attention.

It is **narrow**. Every resampled average lands close to the original, which tells
you the sample mean is a stable, trustworthy estimate.

It is **bell-shaped** - even though the transaction totals themselves were clearly
right-skewed. That is not a coincidence and not specific to this dataset. It is the
Central Limit Theorem, which you meet properly in section 3.2, and it is the reason
most of inferential statistics works at all.

You have just done statistical inference without a single formula. Section 3 shows
you the formulas, and they will make more sense for having seen this first.

# Your turn - full exercises

These have less scaffolding. Work them out from what is above.

**Exercise 1 - Branch performance report.**
Produce one table showing, for each branch: number of transactions, total revenue,
mean and median transaction value, standard deviation, and coefficient of variation.
Sort by total revenue, highest first.

```{r full_ex1, eval=FALSE}
sales |>
  group_by(______) |>
  summarise(
    transactions  = ____,
    total_revenue = sum(______),
    mean_sale     = round(mean(______), 2),
    median_sale   = round(median(______), 2),
    sd_sale       = round(sd(______), 2),
    cv_percent    = round(sd(______) / mean(______) * 100, 1)
  ) |>
  arrange(desc(______))
```

**Exercise 2 - Product line profitability.**
Which product line generates the most total gross income? Is it also the one with
the highest average gross income per transaction? Explain what a difference between
those two answers would mean for the business.

```{r full_ex2, eval=FALSE}
# YOUR CODE HERE
```

**Exercise 3 - Satisfaction by payment method.**
Compare the distribution of `Rating` across the three payment methods, using both a
summary table and a boxplot. Do the differences look large enough to matter?

```{r full_ex3, eval=FALSE}
# YOUR CODE HERE
```

**Exercise 4 - Shape investigation.**
Choose `Unit price`. Calculate its mean, median, skewness and kurtosis, then draw a
histogram and a Q-Q plot. Write two or three sentences describing its shape and
saying whether the mean or the median is the better summary.

```{r full_ex4, eval=FALSE}
# YOUR CODE HERE
```

**Exercise 5 - A question of your own.**
Look at the columns available and ask a question this dataset can answer that has
not been asked above. Answer it with a table and one graph, and write down what a
manager should do differently as a result.

```{r full_ex5, eval=FALSE}
# YOUR CODE HERE
```

Solutions are in `Descriptive-Statistics-SOLUTIONS.Rmd`. Attempt every exercise
before opening it - reading a solution feels like learning and mostly is not.

# Reference card

| Code | What it does |
|---|---|
| `count(d, col)` | Frequency table |
| `count(d, col, sort = TRUE)` | Frequency table, largest first |
| `pivot_wider(names_from = , values_from = )` | Long list into a cross-tab grid |
| `cut(x, breaks = , labels = )` | Group a numeric variable into bands |
| `mean(x)`, `median(x)` | Central tendency |
| `range(x)`, `var(x)`, `sd(x)` | Spread |
| `sd(x) / mean(x) * 100` | Coefficient of variation |
| `quantile(x)` | Five-number summary |
| `quantile(x, probs = c(0.1, 0.9))` | Specific percentiles |
| `IQR(x)` | Interquartile range |
| `skewness(x)`, `kurtosis(x)` | Shape (from `moments`) |
| `cov(x, y)`, `cor(x, y)` | Association |
| `select(d, a, b) |> cor()` | Correlation matrix |
| `ggplot(d, aes(x = )) + geom_bar()` | Bar chart |
| `+ geom_histogram(bins = 30)` | Histogram |
| `+ geom_boxplot()` | Boxplot |
| `+ geom_point(alpha = 0.4)` | Scatterplot |
| `+ stat_qq() + stat_qq_line()` | Q-Q plot |
| `+ coord_flip()` | Turn a chart sideways |
| `+ labs(title = , x = , y = )` | Titles and labels |

# Summary

- **Measurement level** decides which statistics are legitimate. R checks types; only you can check sense.
- **Frequency tables** are the complete analysis for nominal data, and are more informative as percentages.
- **Graphs** reveal what summary numbers hide. Bin width and axis choices are arguments, not neutral defaults.
- **Mean, median, mode** answer "what is typical?" differently. The median is robust; the mean is not.
- **Standard deviation** measures typical distance from the centre. The **coefficient of variation** makes spread comparable across different units.
- **Quartiles and the IQR** give a robust picture of spread and a rule for flagging outliers - but flagging is not deleting.
- **Skewness and kurtosis** describe shape; a **Q-Q plot** shows it far more honestly than two numbers can.
- **Correlation** measures straight-line association only, and never establishes causation. Always plot before you trust it.
- A **bootstrap** turns a single sample into a range of plausible values, which is where section 3 begins.

**Next:** section 3.1, `3-Inferential-Statistics/`, where you stop describing the
data you have and start modelling the process that produced it.
